---
title: "Impact of age and resistance exercise training on intron retention"
format:
  docx:
   reference-doc: resources/custom_reference.docx
author:
  - "Chidimma Phoebe Echebiri^a,b^, Kristian Lian^b^, Torkil Rognefl&aring;ten^b^, 
 H&aring;vard Hamarsland^b^, Knut Sindre M&#248;lmen^b^, Rafi Ahmad^a^, Stian Ellefsen^b^, and Daniel Hammarstr&ouml;m^b^"
editor: visual
csl: resources/embo_reports.csl
bibliography: resources/My_Library.bib
target_journal: https://www.embopress.org/page/journal/14693178/authorguide
prefer-html: true
---

::: {custom-style="author-meta"}
^a^ Department of Biotechnology, University of Inland Norway, Faculty of Applied Ecology, Agricultural Sciences and Biotechnology, Hamar, Norway

^b^ Section for Health and Exercise Physiology, Department of Public Health and Sport Sciences, University of Inland Norway, Lillehammer, Norway

Correspondance: Chidimma Phoebe Echebiri (chidimma.echebiri\@inn.no)
:::

```{r}
#| label: setup
#| include: false


library(dplyr)
library(tidyverse)
library(gt)
library(cowplot)
library(ggpubr)
library(ggplot2)
library(gridExtra)
library(knitr)

color_scale <- c("#7fc97f",  "#beaed4","#fdc086","#ffff99","#386cb0","#f0027f")



source("R/Trainome_functions.R")

knitr::opts_chunk$set(warning = FALSE, message = FALSE)

```

::: {custom-style="abstracttitle"}
Abstract
:::

<div>

</div>

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: abstract

#source("R/Data_exploration.R")
  
  pre_metadata <- readRDS("data_new/Pre_Exercise/all_prexercise_metadata.RDS")
  
  participants <- length(unique(pre_metadata$participant))
  
  desc_stat  <- pre_metadata %>%
    distinct(sex, age, participant) %>%
  
    summarise(.by = c(sex),
              n= n(),
              range = range(age)) %>%
    data.frame()
  
  females <- glue::glue(round(desc_stat[1,2], 0),
                        " females aged ", round(desc_stat[1,3], 0),
                        " to ", round(desc_stat[2, 3], 0))
  
  males <- glue::glue( round(desc_stat[3,2], 0),
                        " males aged ", round(desc_stat[3,3], 0),
                      " to ", round(desc_stat[4, 3], 0))
  
  
  
  full_metadata <- readRDS("data_new/processed_data/all_full_metadata.RDS")
  full_part <- length(unique(full_metadata$participant))
  
  desc_stat_full  <- full_metadata %>%
    distinct(sex, age, participant) %>%
  
    summarise(.by = c(sex),
              n= n(),
              range = range(age)) %>%
    data.frame()
  
  females_full <- glue::glue(round(desc_stat_full[1,2], 0),
                        " females aged ", round(desc_stat_full[1,3], 0),
                        " to ", round(desc_stat_full[2, 3], 0))
  
  males_full <- glue::glue( round(desc_stat_full[3,2], 0),
                        " males aged ", round(desc_stat_full[3,3], 0),
                      " to ", round(desc_stat_full[4, 3], 0))
  
```

Increased intron retention has been linked to aging and several disease conditions, at a rate that is cell and tissue-specific. Investigating factors that influence intron retention would contribute towards understanding gene regulation and the potential mechanisms of reversing or limiting deleterious effects of aging and related disease conditions . This study investigated intron retention and aging in the *vastus lateralis* muscle of `r participants` participants ( `r females` ; `r males`) ; and how it is affected by short-term resistance exercise training in the `r  full_part` participants ( `r females_full` ; `r males_full`) who completed 10 to 13 weeks of training. Using SpliceQ, splicing efficiency of individual introns in paired_end RNA sequence data from six different RT intervention studies was quantified. Differentially spliced introns before the commencement of the exercise intervention studies and at post-intervention were identified using betabinomial generalized linear mixed models (GLMM) . Our results show that aging is associated with increased intron retention in the skeletal muscle. Our results also show that short-term resistance exercise training potentially improves aging-related declines in splicing efficiency. The results contribute towards understanding the adaptive responses to resistance exercise training and highlight its potential for combating aging and aging-related diseases of the skeletal muscle.

# Introduction

Aging is characterized by a progressive decline in most biological functions that affect morbidity and mortality ; [@lopez-otin_hallmarks_2013; @keshavarz_targeting_2023 ; @gyenis_genome-wide_2023; @schaum_ageing_2020] and is associated with the activation of different stress response pathways in several cells and tissues [@stegeman_transcriptional_2017; @welle_gene_2003]. Aging and aging-like phenotypes are also associated with increased mis-regulated alternative splicing [@angarola_splicing_2021 ; @garcia-ruiz_splicing_2025 ; @nishida_tissue-_2015 ; @mariotti_deterioration_2022; @stegeman_transcriptional_2017; @pabis_concerted_2024; @latorre_splicing_2017; @holly_changes_2013] as a result of alterations in the expression of splicing factors . These alterations are resilience strategies aimed at counteracting aging-induced damages and loss of function [@ferrucci_energysplicing_2022 ; @ubaida-mohien_discovery_2019].

Intron retention (IR) is the major type of alternative splicing that is affected by aging [@kodama_systematic_2024; @mariotti_deterioration_2022; @bhadra_alternative_2020], such that increasing IR has been described as a signature of the aging process [@bhadra_alternative_2020; @adusumalli_increased_2019; @baralle_age-related_2023]. IR occurs when a full intron, or parts of an intron in a mature messenger RNA (mRNA) is retained during splicing, resulting in the presence of an unprocessed sequence in the mRNA. This leads either to the degradation of the IR-containing transcripts through nonsense-mediated decay or the action of nucleases; or to frameshifts that cause the generation of alternative protein isoforms [@baralle_age-related_2023 ; @zheng_intron_2020 ; @mauger_targeted_2016; @wong_intron_2022 ; @monteuuis_changing_2019; @jacob_intron_2017 ]. IR mostly affects protein-coding genes, thus resulting in a transcriptomic and proteomic diversity that could influence cellular plasticity. IR can be measured in terms of splicing efficiency (SE) which describes the amount of introns retained in the intron-exon junction of a mature mRNA. Besides aging, poor SE has also been linked to human diseases like alpha thalassemia [@nelson_novel_2005] , inflammatory bowel disease [@hasler_alterations_2011] , neurodegenerative diseases [@sznajder_intron_2018; @ong_increased_2020 ] and an increased risk of certain cancers [@dvinge_widespread_2015 ; @inoue_minor_2021 ; @shah_towards_2022]. It however remains unclear if poor SE is a symptom of the linked conditions , or a mechanism through which the body reduces the impacts of the conditions, or perhaps a causative factor in those conditions. Basically, the exact role of IR in different disease conditions remain to be understood. Some studies [@wang_targeting_2022 ; @naro_splicing_2024 ; @noble_cd19_2023 ] suggest IR could be a therapeutic strategy while others like @jung_intron_2015 suggest it could be a causative factor in diseases or their prognosis. These suggest the role of IR can be good or bad and the challenge potentially lies in differentiating pathological IR from therapeutic , regulatory or symptomatic IR.

Progressive resistance exercise training (RT) is a type of high intensity training primarily aimed at the skeletal muscle system but with system-wide benefits that include improvements in muscle strength and mass, increased bone density, glucose homeostasis and several other hallmarks of aging [@garatachea_exercise_2015; @liu_progressive_2009]. RT could thus be a potent preventive or treatment strategy for improving functional disabilities associated with the aging skeletal muscle [@peterson_influence_2011 ; @kryger_resistance_2007 ; @lu_effects_2021 ], reducing all-cause mortality [@ekelund_dose-response_2019] and improving cardiorespiratory fitness in older persons [@smart_role_2022]. A lot is currently known about adaptive responses to RT. However, intron retention in response to RT in healthy individuals has largely been unexplored except by few studies like @kodama_systematic_2024 . In this study, we investigated the relationship between aging and IR in human skeletal muscle using RNA sequences from the vastus lateralis muscle of `r participants` participants ( `r females` ; `r males`) from six RT intervention studies . We also investigated how RT affected IR in the muscle of `r  full_part` participants ( `r females_full` ; `r males_full`) who performed 10 to 13 weeks of RT . Here, we show that SE declines with aging , and that RT could improve age-related decline in SE. The characteristics of included studies are as shown in **Table 1**. Figure 1 provides a summarised description of how the study was conducted.

Table 1: Basic characteristics of RT intervention studies included in this analysis: primary results from three of the RT intervention studies (COPD, Volume; SRP102542) have been previously published. Results from ContraTrain, Alpha & Omega, and ReLiEf are yet to be published but their clinical identifier numbers are provided

+------------------------------------+------------------------+----------------------+------------+--------------------------------+
| Study                              | Number of participants | Age Range            | Duration   | Publication/Clinical identifer |
+====================================+========================+======================+============+================================+
| Alpha/Omega                        | 143                    | 23-60                | 12         | NCT04279951                    |
+------------------------------------+------------------------+----------------------+------------+--------------------------------+
| ContraTrain                        | 32                     | 19-35                | 10         | NCT03795025                    |
+------------------------------------+------------------------+----------------------+------------+--------------------------------+
| COPD                               | 53                     | 56-79                | 13         | @molmen_chronic_2021           |
+------------------------------------+------------------------+----------------------+------------+--------------------------------+
| ReLief                             | 66                     | 20-29 (Young cohort) | 12         | NCT05063279                    |
|                                    |                        |                      |            |                                |
|                                    |                        | 72-93(Old cohort)    |            |                                |
+------------------------------------+------------------------+----------------------+------------+--------------------------------+
| SRP102542 (GEO Accession GSE97084) | 52                     | 19-30 (Young cohort) | 12         | @robinson_enhanced_2017        |
|                                    |                        |                      |            |                                |
|                                    |                        | 65-78 (Old cohort)   |            |                                |
+------------------------------------+------------------------+----------------------+------------+--------------------------------+
| Volume                             | 25                     | 20-37                | 12         | @khan_increased_2020           |
+------------------------------------+------------------------+----------------------+------------+--------------------------------+

# Results and Discussion

## Intron retention  is associated with aging in the skeletal muscle

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: Results_1

# The full spliceQ baseline data

ds_introns_pre <- readRDS("data_new/baseline_model_outputs/baseline_model_df.RDS")

# Load the gene annotation file
 gene_annotation <- readRDS("data_new/ensembl_gene_annotation.RDS")

ds <- length(unique(ds_introns_pre$target))

ds_pos_ <- ds_introns_pre %>%
   filter(Estimate > 0 & Pr...z.. <= 0.05)
 ds_pos <-  length(unique(ds_pos_$target))


ds_neg_ <- ds_introns_pre %>%
   filter(Estimate < 0 & Pr...z.. <= 0.05) 

ds_neg <- length(unique(ds_neg_$target))


ds_total <- ds_neg + ds_pos
# EXplore the gene and transcript level information of the ds introns

# How many transcripts in full data
transcript_level <- length(unique(ds_introns_pre$transcript_ID))

gene_level <- length(unique(ds_introns_pre$ensembl_gene_id))



```

To investigate the relationship between IR and aging in the human *vastus lateralis* muscle, we analysed the SE scores of `r ds` introns based on the inclusion criteria that samples were taken from participants before commencement of the exercise intervention, that the introns were quantified across all samples and that they did not have an SE score of one across all samples. there was no obvious association between age-related IR and the chromosomal location of the genes or their gene expression counts

Splicing efficiency is said to be cell, tissue and specie specific . Previous studies [@garcia-ruiz_splicing_2025; @baralle_age-related_2023; @mariotti_deterioration_2022; @bhadra_alternative_2020; @adusumalli_increased_2019] had linked poor SE to aging in different tissues . These introns were from `r transcript_level` transcripts of `r gene_level` genes. SpliceQ @de_melo_costa_splice-q_2021 was used to quantify SE . The tool awards each intron an SE score ranging from zero to one. A score of zero means that the intron is fully retained, while a score of one means the complete splicing-out of the intron. Scores between zero and one indicate the approximate fraction of retained nucleotides.

Of the `r ds` introns , `r ds_total` were significantly affected by aging; the SE scores of `r ds_pos` introns improved with aging , but declined in `r ds_neg` introns . Figure 2A shows the SE trend with aging in both the introns affected by aging and those not affected. Our results indicate that while the SE scores of most introns in the skeletal muscle are unaffected by aging, the scores of a relevant number of them are changed by aging, mostly in form of declining SE scores with age.

We further explored the transcript and gene level information of the introns to determine if there existed a pattern . To achieve that, we grouped the introns into those not affected by aging (No effect), those with SE scores that improved with aging ( Improved SE), and those with age-related decline in their SE scores (Reduced SE). Our results show that the effects of aging are intron-specific as several transcripts contained introns that cut across the several groups as shown in Figure 2B. 61 transcripts contained introns unaffected by aging, those whose SE scores were reduced by aging, as well as those improved by aging. There were transcripts that intersected between the different groups. This result agrees with those obtained by @sun_abnormalities_2023 that suggests differential regulation of splicing at the intron-level in addition to multiple other levels [@monteuuis_changing_2019]. We also observed no gene or transcript-level expression patterns among the different groups. This is not surprising given the intron-specificity of the effects.

Majority of the introns in all three groups were contained in protein-coding genes. Previous studies suggest IR occurs mostly in protein-coding genes, but we do not know how much of that is attributable to short-read RNA sequencing which tends to be biased towards quantification of protein-coding genes. Similar to @stokes_transcriptomics_2023 contesting the long-held belief that long non-coding RNAs are sparsly expressed in comparison to protein-coding genes, it might be necessary to explore IR using alternative RNA methods to confirm its occurrence in mainly protein-coding genes and investigate if truly short-read RNA sequencing is underestimating IR as suggested by @garcia-ruiz_splicing_2025 .

In addition, we explored the functional annotation of the genes containing introns affected by aging, Gene ontology (GO) analyses revealed that the introns affected by aging are involved mostly in muscle-related biological processes as can be shown in Figure 2C. This would not be unrelated to the known effects of aging on the skeletal muscle where functions decline with increasing age.

## Resistance exercise training affects splicing efficiency and could improve aging-related declines in splicing efficiency

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: Results_4

all_pre_metadata <- readRDS("data_new/Pre_Exercise/all_prexercise_metadata.RDS")

all_full_metadata <- readRDS("data_new/processed_data/all_full_metadata.RDS")


pre_samples <- length(unique(all_pre_metadata$seq_sample_id))

full_samples <- length(unique(all_full_metadata$seq_sample_id))

all_post_splice <- readRDS("data_new/processed_data/all_splice_data.RDS")

all_splice <- length(unique(all_post_splice$transcript_ID))

# Load the baseline model output to query intersects between that and RT
baseline_df <- readRDS("data_new/baeline_merged_model_output.RDS") %>%
  filter(effect != "No effect")
# 



RT_effects <- readRDS("data_new/RT_model_outputs/RT_model_df.RDS")%>%
  filter(effect != "No effect" )

 RT_introns <- length(unique(RT_effects$target))
 
 RT_transcripts <- length(unique(RT_effects$transcript_ID))
 
 RT_genes <- length(unique(RT_effects$ensembl_gene_id))
 
 filt_exc <- RT_effects %>%
  filter(coef == "timePostExc")
 
 filt_int <- RT_effects %>%
  filter(coef == "scaled_age:timePostExc")
 
exc_alone <- length(unique(filt_exc$target))

int_both <- length(unique(filt_int$target))

int_base <-  RT_effects %>%
  filter( target %in% baseline_df$target)

RT_and_base <- length(unique(int_base$target))

Improved_RT <- RT_effects %>%
  filter(effect == "Improved SE")


Reduced_RT <- RT_effects %>%
  filter(effect == "Reduced SE")


Inte_both <- length(intersect(Improved_RT$transcript_ID, Reduced_RT$transcript_ID))
```

Having established that aging affects splicing efficiency in human skeletal muscle, we sought to establish if and how much short term (\~10-13 weeks) RT could impact age-related decline in SE, and SE in general. To achieve this objective, we filtered our data to include only participants who performed RT. Thus, the control samples from participants in the ContraTrain (NCT03795025) and ReLiEf (NCT05063279) studies , as well as the participants who performed non-RT interventions in @robinson_enhanced_2017 were excluded from this step. The number of samples however increased to `r full_samples` as both pre-exercise and post-exercise samples were included in this step. Filtering for introns represented across all samples and excluding those with scores of one across all samples reduced the number of introns to 8368 . `r RT_introns` introns from `r RT_transcripts` transcripts of `r RT_genes` genes were differentially spliced due to the impact of RT, either alone (`r exc_alone`) or in interaction with age (`r int_both`). Though more introns were differentially spliced due to the impact of RT alone, its interaction with aging had more effect on intron retention as 248 introns had LogF2C values at or above 0.5. That is, there was at least a doubling of the impact on SE due to the interaction of RT and aging in 248 introns, as compared to 8 due to RT alone. Like the impact of aging, the impact of RT was also intron-specific.

To understand how RT affected introns that had been confirmed to be affected by aging in the first model, we filtered for introns with aging-related changes in splicing efficiency that were also affected by RT. Figure 3A shows the SE trend of the `r RT_and_base` introns affected by aging and RT. RT affected SE in more that 10% of the introns in the dataset. GO analyses showed these were introns in genes mainly involved in muscoskeletal processes as shown in Figure 3B.

Exercise is already known to counteract the effects of aging and can offset aging-related changes to the splicing machinery @kodama_systematic_2024, @ubaida-mohien_physical_2019 and @ubaida-mohien_unbiased_2022 , but we contribute to what is already known by focusing mainly on RT and intron retention. The pattern of response to both aging and RT point towards strong regulatory roles of IR and there is need to experimentally explore IR in response to RT as this would further deepen our insights into transcriptomic responses to RT.

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: Results


all_pre_splice <- readRDS("data_new/Pre_Exercise/all_pre_Exc_splicing_data.RDS")

introns <- length(unique(all_pre_splice$transcript_ID))

all_post_splice <- readRDS("data_new/processed_data/PostEXc_splice_data.RDS")
all_post_introns <- length(unique(all_post_splice$transcript_ID))

splice_int <- setdiff(all_post_splice$transcript_ID, all_pre_splice$transcript_ID)

# Load the poorly spliced introns table
low_SE <- readRDS("data_new/Pre_Exercise/annotated_low_SE_introns.RDS") %>%
  dplyr::select(transcript_ID, intron_ID, transcript_biotype,external_gene_name, min,mode, max)
kable(low_SE, caption = " Table 1:Introns with poor splicing efficiency across all baseline samples")


perfect_SE <- readRDS("data_new/Pre_Exercise/annotated_perfect_SE_introns.RDS")

perfect_SE <- perfect_SE %>%
  # extract the full_intron_id by combining the transcript name and intronid
  mutate(full_intron_Id = paste0(perfect_SE$transcript_ID, "_", perfect_SE$intron_ID, "_", perfect_SE$chr))

PE_introns <- length(perfect_SE$intron_ID)

PE_transcripts <- length(unique(perfect_SE$transcript_ID))

PE_genes <- length(unique(perfect_SE$ensembl_gene_id))

perfect_SE_post <- readRDS("data_new/processed_data/annotation_High_SE_postEXc.RDS")


perfect_SE_post <- perfect_SE_post %>%# extract the full_intron_id by combining the transcript name and intronid
  mutate(full_intron_Id = paste0(perfect_SE_post$transcript_ID, "_", perfect_SE_post$intron_ID, "_",perfect_SE_post$chr ))


# which introns were perfectly spliced in post but werent at pre

post_not_pre <- setdiff(perfect_SE_post$full_intron_Id, perfect_SE$full_intron_Id)
 # were they in the baseline data?
intersect(all_pre_splice$transcript_ID, post_not_pre)


post_introns <- length(perfect_SE_post$full_intron_Id)

post_transcripts <- length(unique(perfect_SE_post$transcript_ID))

post_genes <- length(unique(perfect_SE_post$ensembl_gene_id))


no_dev_pre <- readRDS( "data_new/processed_data/annotation_no_deviation_PreExc.RDS")


no_dev_transcripts<- length(unique(no_dev_pre$transcript_ID))

no_dev_introns <- length(no_dev_pre$transcript_ID)

no_dev_genes <- length(unique(no_dev_pre$ensembl_gene_id))




```

## Splicing efficiency may be related to physiological relevance of a transcript but not to gene expression values

Previous research report a correlation between SE of an intron and the physiological relevance of the transcript containing the intron [@braunschweig_widespread_2014 ; @saudemont_fitness_2017; @sanchez-escabias_co-transcriptional_2022]. We investigated if this applies to the skeletal muscle by identifying introns with very low SE, as well as those with perfect SE across all samples . To identify those with low SE, we filtered the pre-exercise data for introns with a maximum total SE of 0.2 across all samples . Six introns belonging to six different transcripts were shown to be almost completely retained in all samples. Five of these transcripts are long non-coding RNAs without functional annotation on the org.Hs.eg.db database [@manual_orghsegdb_nodate] , while the sixth is an alternatively spliced transcript of the MEIOC gene that lacks a coding sequence . The introns and their characteristics are as shown in **Table 2.** GO analysis of MEIOC reveals it is involved in meiotic processes.

Furthermore, we filtered for introns that were completely spliced out across all samples. `r PE_introns` introns from `r PE_transcripts` transcripts belonging to `r PE_genes` genes were perfectly spliced across all the pre-exercise samples. These genes were mostly protein_coding genes enriched for biological processes that are relevant to the skeletal muscle as can be seen in **Figure 4B**. Like previously mentioned, the occurrence of IR in mostly protein-coding genes may just be a consequence of short-read RNA sequencing as even the transcripts containing introns with perfect SE scores were predominantly in protein-coding genes.

Of the `r all_post_introns` introns recorded in the postexercise data, there were `r post_introns` perfectly spliced introns belonging to `r post_transcripts` transcripts of `r post_genes` genes. Of these introns, 50 were among those containing retained introns at baseline. Again, only two introns were completely retained in the post-exercise data, both of which were among the non-functionally annotated lncRNAs also retained in the pre-exercise data. These also highlight RT's potential at improving SE. Given the numbers of introns captured in the pre-exercise and the post-exercise data, basic data analyses revealed that 1 % of the introns were completely spliced out in the former while 2.5% of the latter were completely spliced out; including those with retained introns at pre-exercise. This suggests an improvement in SE attributable to RT. These perfectly spliced out introns were shown to be physiologically relevant to the skeletal muscle, thus, supporting the presence of a relationship between SE and relevance in a cell or tissue [@sanchez-escabias_co-transcriptional_2022; @saudemont_fitness_2017; @braunschweig_widespread_2014] . However, the intron-specificity of the effects of both aging and RT questions that notion as different introns in the same transcripts could be responding differently to interventions or conditions. This clearly suggests that physiological relevance alone does not determine SE as both internal factors like aging, and response to external factors like RT potentially determines SE. Like most biological processes, there would likely be an intricate web of predictive variables in the relationship between physiological relevance and SE.

## Splicing efficiency of introns in a gene tend to be similar

According to @sanchez-escabias_co-transcriptional_2022, SE tends to be similar between the introns of a gene . We investigated the presence of that similarity in our dataset by filtering for introns with standard deviation of 0.01 across all pre-exercise samples. Our results show that `r no_dev_introns` introns in `r no_dev_transcripts` transcripts of `r no_dev_genes` had similar SE across all samples. This suggests that more than half of the introns were similarly spliced . The number of similarly spliced introns at the transcript and gene levels ranged from 1 to 45 and 1 to 311 respectively; this is as compared to the entire introns in the pre-exercise dataset which ranged from 1 to 450 at the gene level and 1 to 58 at the transcript level . These suggest that most introns in a transcript or gene tend to have similar splicing efficiency. Evident from our results is the fact that physiological factors such as aging, and external factors can disrupt that similarity.

## Transcript length does not correlate to splicing efficiency in the skeletal muscle

There is said to exist a positive correlation between gene length and SE [@sanchez-escabias_co-transcriptional_2022 ; @saudemont_fitness_2017; @braunschweig_widespread_2014]. We investigated this by using Pearson's product-moment correlation to check for correlations between transcript length and SE of introns in a transcript. A negligible correlation of -0.018 at a pvalue of 0.007 was found between the transcript length and SE of the samples at baseline. When tested in the post exercise data alone , they had a correlation estimate of -0.054 at a p value of 0.0. In the full dataset (baseline and postexercise), the correlation was -0.042 with a p.value of 0.0. All the results completely disagree with reports that suggest a relationship between those two factors in human skeletal muscles.

# Limitations

This study did not group RT into conditions or volume of exercise interventions . The benefits of RT are known to be volume-dependent and the 6 studies involved varying RT conditions. We were thus unable to explore which RT conditions would offer the optimal SE outcomes.

# Materials and methods

## Datasets

Six different studies designed to investigate different aspects of adaptive responses to RT were included in the analyses. Five of these studies were conducted in our research group, while one is a study by @robinson_enhanced_2017 with GEO ascension number GSE97084. The studies were selected based on the criteria that they included paired-end Illumina RNA-Seq data from the *vastus lateralis* muscle of human subjects who performed at least 10 weeks of RT with samples collected at baseline and at the end of the study . Age of all participants was also required to be included in the data. The study protocols of the different studies are described elsewhere. A flowchart of the data analyses is as seen in **Figure 1** . **Figure 4** shows the distribution of participants within the individual studies.

## RNA sequence data processing

STAR (2.7.9a) [@dobin_star_2013] was used to map the RNA Sequence reads to the human reference genome (GRCh38, gencode v47 primary annotation ) using the 2-pass alignment method described by @veeneman_two-pass_2016. This method is said to advance the quantification and discovery of splicing events. RSEM (v.1.3.3-foss-2019b) @li_rsem_2011 was used to quantify transcript and gene level expressions. Splicing efficiency of each intron was quantified using the default settings in SpliceQ [@de_melo_costa_splice-q_2021] whereby introns are scored only if they have a minimum intron length of 30 nucleotides, have a minimum of 10 reads spanning each splice junction, above 10 uniquely mapping reads, and if they do not overlap with any exon; among other settings .

## Data analyses

Analyses of the extracted SpliceQ data for all participants in the six studies were done using R version 4.4.1 on RStudio 2024.04.2+764 "Chocolate Cosmos" Release . Introns included in each data analyses step were only those whose SE was quantified across all samples.

## Investigating the impact of aging , and RT on splicing efficiency

To examine the effect of aging on splicing efficiency , we transformed participants' age to a scale of 0 to 1 and fitted a generalized linear mixed model (GLMM) using a Beta distribution. The model built using only the pre-exercise data included random intercepts for study and participants to account for random variations at the study and participant levels while sex and scaled age of participants were fixed effects. . Differentially spliced introns were defined as those with p values at or below 0.05.

To examine the effect of \~12 weeks RT on SE, we used a similar model that included fixed interactive effects for aging and RT, and sex. Study and participants remained random effects . Only samples that performed RT were included in this analyses. Thus, the control samples in ContraTrain, RELiEf and the non-RT samples in @robinson_enhanced_2017 were not included in the second model.

# References
