---
title: "Impact of age and resistance exercise training on intron retention"
format:
  docx:
   reference-doc: resources/custom_reference.docx
author:
  - "Chidimma Phoebe Echebiri^1,2^, Kristian Lian^2^, Torkil Rognefl&aring;ten^2^, 
 H&aring;vard Hamarsland^2^, Knut Sindre M&#248;lmen^2^, Rafi Ahmad^13^, Stian Ellefsen^2^, and Daniel Hammarstr&ouml;m^2^"
editor: visual
csl: resources/molecular_systems_biology.csl
bibliography: resources/My_Library.bib
target_journal: https://www.embopress.org/page/journal/17444292/authorguide
prefer-html: true
---

::: {custom-style="author-meta"}
1 Department of Biotechnology, University of Inland Norway, Faculty of Applied Ecology, Agricultural Sciences and Biotechnology, Holsetgata 31, 2318 Hamar, Norway

2 Section for Health and Exercise Physiology, Department of Public Health and Sport Sciences, University of Inland Norway, Vormstuguvegen 2, 2624 Lillehammer, Norway

3 Institute for Clinical Medicine, Faculty of Health Sciences, UiT- The Arctic University of Norway, Hansine Hansens veg 18, 9019, Tromsř, Norway

Correspondence: Chidimma Phoebe Echebiri (chidimma.echebiri\@inn.no)
:::

```{r}
#| label: setup
#| include: false


library(dplyr)
library(tidyverse)
library(gt)
library(cowplot)
library(ggpubr)
library(ggplot2)
library(gridExtra)
library(knitr)

color_scale <- c("#7fc97f",  "#beaed4","#fdc086","#ffff99","#386cb0","#f0027f")



source("R/Trainome_functions.R")

knitr::opts_chunk$set(warning = FALSE, message = FALSE)

```

::: {custom-style="abstracttitle"}
Abstract
:::

<div>

</div>

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: abstract

#source("R/Data_exploration.R")
  
  pre_metadata <- readRDS("data_new/Pre_Exercise/all_prexercise_metadata.RDS")
  
  participants <- length(unique(pre_metadata$participant))
  
  desc_stat  <- pre_metadata %>%
    distinct(sex, age, participant) %>%
  
    summarise(.by = c(sex),
              n= n(),
              range = range(age)) %>%
    data.frame()
  
  females <- glue::glue(round(desc_stat[1,2], 0),
                        " females aged ", round(desc_stat[1,3], 0),
                        " to ", round(desc_stat[2, 3], 0))
  
  males <- glue::glue( round(desc_stat[3,2], 0),
                        " males aged ", round(desc_stat[3,3], 0),
                      " to ", round(desc_stat[4, 3], 0))
  
  
  
  full_metadata <- readRDS("data_new/processed_data/all_full_metadata.RDS")
  full_part <- length(unique(full_metadata$participant))
  
  desc_stat_full  <- full_metadata %>%
    distinct(sex, age, participant) %>%
  
    summarise(.by = c(sex),
              n= n(),
              range = range(age)) %>%
    data.frame()
  
  females_full <- glue::glue(round(desc_stat_full[1,2], 0),
                        " females aged ", round(desc_stat_full[1,3], 0),
                        " to ", round(desc_stat_full[2, 3], 0))
  
  males_full <- glue::glue( round(desc_stat_full[3,2], 0),
                        " males aged ", round(desc_stat_full[3,3], 0),
                      " to ", round(desc_stat_full[4, 3], 0))
  
```

Increased intron retention has been linked to aging and several disease conditions, at a rate that is cell and tissue-specific. Investigating factors that influence intron retention would contribute towards understanding gene regulation and the potential mechanisms of reversing or limiting deleterious effects of aging and related disease conditions . This study investigated intron retention and aging in the *vastus lateralis* muscle of `r participants` participants ( `r females` ; `r males`) ; and how it is affected by short-term resistance exercise training in the `r  full_part` participants ( `r females_full` ; `r males_full`) who completed 10 to 13 weeks of training. Using SpliceQ, splicing efficiency of individual introns in paired_end RNA sequence data from six different RT intervention studies was quantified. Differentially spliced introns before the commencement of the exercise intervention studies and at post-intervention were identified using betabinomial generalized linear mixed models (GLMM) . Our results show that aging is associated with increased intron retention in the skeletal muscle. Our results also show that short-term resistance exercise training potentially improves aging-related declines in splicing efficiency. The results contribute towards understanding the adaptive responses to resistance exercise training and highlight its potential for combating aging and aging-related diseases of the skeletal muscle.

# Introduction

Aging is characterized by a progressive decline in most biological functions that affect morbidity and mortality ; [@lopez-otin_hallmarks_2013; @keshavarz_targeting_2023 ; @gyenis_genome-wide_2023; @schaum_ageing_2020] and is associated with the activation of different stress response pathways in several cells and tissues [@stegeman_transcriptional_2017; @welle_gene_2003]. Aging and aging-like phenotypes are also associated with increased mis-regulated alternative splicing [@angarola_splicing_2021 ; @garcia-ruiz_splicing_2025 ; @nishida_tissue-_2015 ; @mariotti_deterioration_2022; @stegeman_transcriptional_2017; @pabis_concerted_2024; @latorre_splicing_2017; @holly_changes_2013] as a result of alterations in the expression of splicing factors . These alterations are resilience strategies aimed at counteracting aging-induced damages and loss of function [@ferrucci_energysplicing_2022 ; @ubaida-mohien_discovery_2019].

Intron retention (IR) is the major type of alternative splicing that is affected by aging [@kodama_systematic_2024; @mariotti_deterioration_2022; @bhadra_alternative_2020], such that increasing IR has been described as a signature of the aging process [@bhadra_alternative_2020; @adusumalli_increased_2019; @baralle_age-related_2023]. IR occurs when a full intron, or parts of an intron in a mature messenger RNA (mRNA) is retained during splicing, resulting in the presence of an unprocessed sequence in the mRNA. This leads either to the degradation of the IR-containing transcripts through nonsense-mediated decay or the action of nucleases; or to frameshifts that cause the generation of alternative protein isoforms [@baralle_age-related_2023 ; @zheng_intron_2020 ; @mauger_targeted_2016; @wong_intron_2022 ; @monteuuis_changing_2019; @jacob_intron_2017 ]. IR mostly affects protein-coding genes, thus resulting in a transcriptomic and proteomic diversity that could influence cellular plasticity. IR can be measured in terms of splicing efficiency (SE) which describes the amount of introns retained in the intron-exon junction of a mature mRNA. Besides aging, poor SE has also been linked to human diseases like alpha thalassemia [@nelson_novel_2005] , inflammatory bowel disease [@hasler_alterations_2011] , neurodegenerative diseases [@sznajder_intron_2018; @ong_increased_2020 ] and an increased risk of certain cancers [@dvinge_widespread_2015 ; @inoue_minor_2021 ; @shah_towards_2022]. It however remains unclear if poor SE is a symptom of the linked conditions , or a mechanism through which the body reduces the impacts of the conditions, or perhaps a causative factor in those conditions. Basically, the exact role of IR in different disease conditions remain to be understood. Some studies [@wang_targeting_2022 ; @naro_splicing_2024 ; @noble_cd19_2023 ] suggest IR could be a therapeutic strategy while others like @jung_intron_2015 suggest it could be a causative factor in diseases or their prognosis. These suggest the role of IR can be good or bad and the challenge potentially lies in differentiating pathological IR from therapeutic , regulatory or symptomatic IR.

Progressive resistance exercise training (RT) is a type of high intensity training primarily aimed at the skeletal muscle system but with system-wide benefits that include improvements in muscle strength and mass, increased bone density, glucose homeostasis and several other hallmarks of aging [@garatachea_exercise_2015; @liu_progressive_2009]. RT could thus be a potent preventive or treatment strategy for improving functional disabilities associated with the aging skeletal muscle [@peterson_influence_2011 ; @kryger_resistance_2007 ; @lu_effects_2021 ], reducing all-cause mortality [@ekelund_dose-response_2019] and improving cardiorespiratory fitness in older persons [@smart_role_2022]. A lot is currently known about adaptive responses to RT. However, intron retention in response to RT in healthy individuals has largely been unexplored except by few studies like @kodama_systematic_2024 . In this study, we investigated the relationship between aging and IR in human skeletal muscle using RNA sequences from the vastus lateralis muscle of `r  full_part` participants ( `r females_full` ; `r males_full`) ,and the relationship between aging and 10 to 13 weeks of RT . Here, we show that SE declines with aging , and that RT could improve age-related decline in SE. The characteristics of included studies are as shown in **Table 1**. Figure 1 provides a summarised description of how the study was conducted.

Table 1: Basic characteristics of RT intervention studies included in this analysis: primary results from three of the RT intervention studies (COPD, Volume; SRP102542) have been previously published. Results from ContraTrain, Alpha & Omega, and ReLiEf are yet to be published but their clinical identifier numbers are provided

+------------------------------------+------------------------+----------------------+------------+--------------------------------+
| Study                              | Number of participants | Age Range            | Duration   | Publication/Clinical identifer |
+====================================+========================+======================+============+================================+
| Alpha/Omega                        | 143                    | 23-60                | 12         | NCT04279951                    |
+------------------------------------+------------------------+----------------------+------------+--------------------------------+
| ContraTrain                        | 32                     | 19-35                | 10         | NCT03795025                    |
+------------------------------------+------------------------+----------------------+------------+--------------------------------+
| COPD                               | 53                     | 56-79                | 13         | @molmen_chronic_2021           |
+------------------------------------+------------------------+----------------------+------------+--------------------------------+
| ReLief                             | 66                     | 20-29 (Young cohort) | 12         | NCT05063279                    |
|                                    |                        |                      |            |                                |
|                                    |                        | 72-93(Old cohort)    |            |                                |
+------------------------------------+------------------------+----------------------+------------+--------------------------------+
| SRP102542 (GEO Accession GSE97084) | 52                     | 19-30 (Young cohort) | 12         | @robinson_enhanced_2017        |
|                                    |                        |                      |            |                                |
|                                    |                        | 65-78 (Old cohort)   |            |                                |
+------------------------------------+------------------------+----------------------+------------+--------------------------------+
| Volume                             | 25                     | 20-37                | 12         | @khan_increased_2020           |
+------------------------------------+------------------------+----------------------+------------+--------------------------------+

# Results 

## Splicing efficiency of some introns in the skeletal muscle is associated with aging

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: RT_results

# Load the model summaries for each intron
introns <- readRDS("data_new/RT_model_outputs/RT_model_df.RDS")
length_introns <- length(unique(introns$target))
length_genes <- length(unique(introns$external_gene_name))

# Load splicing data
all_splice_df <- readRDS("data_new/processed_data/all_splice_data.RDS")

High_SE_df <- all_splice_df %>%
  pivot_longer(names_to = "seq_sample_id",
               values_to = "SE",
               cols = -(transcript_ID)) %>%
  summarise(.by = transcript_ID, 
            mode = getmode(SE),
            min = min(SE), 
            max = max(SE)) %>%
  filter(min == 1) 

length_perfect_scores <- length(unique(High_SE_df$transcript_ID))


# Load the merged data that includes the prediction and gene annotation of the model outp



# Extract the effects of aging alone
Aging_effects <- introns %>%
  filter(coef == "scaled_age" & time == "PreExc" & effect != "No effect")

length_age_effects <- length(unique(Aging_effects$target))

RT_effects <- introns %>%
  filter(coef == "timePostExc" & time == "PreExc" & effect != "No effect")
length_RT_introns <- length(unique(RT_effects$target))
```

We investigated the relationship between aging and SE in the human *vastus lateralis* muscle using a Generalised Linear Mixed Model . Of the `r length_introns` quantified across all samples in our studies, the SE scores of 30908 of them were not affected by aging. 2221 introns were affected by aging, of which 512 showed improved SE with aging while 1709 had reduced SE with aging (Fig 2A). There was no apparent association between the genes containing these differentially spliced (ds) introns and their chromosomal locations (Fig 2B). The rate of SE improvements due to aging were relatively smaller compared to the rate of decline as can be seen in Supplementary Figure 1. Most of the introns with reduced SE upon aging were those with SE scores below 0.95 in the youngest participants. It appeared that the decline had probably began before the youngest participant, or that these were introns already prone to poor SE scores. Based on the model's predictions of SE scores with aging, the topmost affected introns were ranked and are shown in supplementary images. These effects were intron-specific as a gene could contain both introns affected by aging, and those not affected by aging as shown in Figure 2C. Gene ontology analyses revealed that the ds introns belong mostly to genes involved in muscle-related biological processes and were mostly located in the myofibril and contractile fiber (Fig 2D and E).

## Resistance exercise training has age-dependent effects on splicing efficiency and could improve aging-related declines in splicing efficiency

SE of `r length_RT_introns` introns were associated with RT. Unlike aging, most of the associations involved improvements in SE scores post RT intervention (Fig 3A). These effects largely seemed to be age-dependent as 1593 of the `r length_RT_introns` introns showed an age-dependent response to RT (Figure 3). Like the impact of aging, the impact of RT was also intron-specific.The genes sets were involved in the same processes as those affected by aging, albeit to differing gene ratios (Fig 3C). 384 introns from 51 genes with age-related changes in SE had slightly improved SE scores following short-term RT

## Splicing efficiency of specific introns in a gene could be same despite transcript type

our results reveal that the SE scores of specific intron junctions in a gene were same, even when in different transcripts . We hypothesize that these are the same intron-exon junctions but due to other types of alternative splicing events, occur in different transcripts of the same gene. See supplementary figure

## Most introns are in protein-coding genes

A good majority of the introns in the entire dataset, as well as those affected by either RT or aging are contained in protein-coding genes

## Some introns are perfectly spliced out across all ages and conditions

201 introns were perfectly spliced (SE score of 1) across all samples. Like the pattern in the full dataset, these introns were contained in mostly protein-coding genes

## Transcript length is not associated with SE

We investigated the relationship between transcript length and SE. There appears to be no relationship between he two

# Discussion

This would not be unrelated to the known effects of aging on the skeletal muscle where functions tends to decline with increasing age. We also investigated the intron-specificity of the effects (or lack thereof) of aging by exploring if the effects were specific to certain introns in certain genes. Our results suggest that the effects of aging are intron-specific as several genes contained introns that cut across the several groups as shown in Figure 2E which suggests differential regulation of splicing at the intron-level in agreement with @sun_abnormalities_2023 in addition to other levels of regulation [@monteuuis_changing_2019]. Majority of the introns in all three groups were contained in protein-coding genes in agreement with previous studies that suggest IR occurrence in mostly protein-coding genes. However, this is arguably attributable to short-read RNA sequencing and its bias towards quantification of protein-coding genes. Similar to @stokes_transcriptomics_2023 contesting the long-held belief that long non-coding RNAs are sparsely expressed in comparison to protein-coding genes, it might be necessary to explore IR using alternative RNA methods to confirm its occurrence in mainly protein-coding genes and investigate if truly short-read RNA sequencing is underestimating IR as suggested by @garcia-ruiz_splicing_2025 .

Though our results agree with previous studies that linked declining SE to aging in different tissues [@garcia-ruiz_splicing_2025; @baralle_age-related_2023; @mariotti_deterioration_2022; @bhadra_alternative_2020; @adusumalli_increased_2019], we however show that this is observed in less than one percent of introns in the skeletal muscle . Also, our results show that while most age-associated effects are declines in SE, SE of some introns potentially improve with age. This suggests that the role of intron retention in gene regulation should not be generalised as bad or good, and that further studies are needed in understanding the specific effects of intron retention in relation to different aging-related conditions.

The relatively small response to RT may be a result of the short-duration and thus may require longer-term RT to elicit more significant impacts on SE. However, the range of changes in response to short-term RT appeared relatively small . This suggests that longer-term RT, or perhaps a lifetime of RT would have a much impacts on aging

Exercise is already known to counteract the effects of aging and can offset aging-related changes to the splicing machinery @kodama_systematic_2024, @ubaida-mohien_physical_2019 and @ubaida-mohien_unbiased_2022 , but we contribute to what is already known by focusing mainly on RT and intron retention. The pattern of response to both aging and RT point towards regulatory roles of IR and there is need to explore IR in response to long-term RT to ascertain if the impact on IR depends on factors like length of RT as well as RT volume and conditions. Thus, there is need to further deepen our insights into transcriptomic responses to RT.

# Limitations

This study did not group RT into conditions or volume of exercise interventions . The benefits of RT are known to be volume-dependent and the 6 studies involved varying RT conditions. We were thus unable to explore which RT conditions would offer the optimal SE outcomes.

# Materials and methods

## Datasets

Six different studies designed to investigate different aspects of adaptive responses to RT were included in the analyses. Five of these studies were conducted in our research group, while one is a study by @robinson_enhanced_2017 with GEO ascension number GSE97084. The studies were selected based on the criteria that they included paired-end Illumina RNA-Seq data from the *vastus lateralis* muscle of human subjects who performed at least 10 weeks of RT with samples collected at baseline and at the end of the study . Age of all participants was also required to be included in the data. The study protocols of the different studies are described elsewhere. A flowchart of the data analyses is as seen in **Figure 1** . **Figure 4** shows the distribution of participants within the individual studies.

## RNA sequence data processing

STAR (2.7.9a) [@dobin_star_2013] was used to map the RNA Sequence reads to the human reference genome (GRCh38, gencode v47 primary annotation ) using the 2-pass alignment method described by @veeneman_two-pass_2016. This method is said to advance the quantification and discovery of splicing events. RSEM (v.1.3.3-foss-2019b) @li_rsem_2011 was used to quantify transcript and gene level expressions. Splicing efficiency of each intron was quantified using the default settings in SpliceQ [@de_melo_costa_splice-q_2021] whereby introns are scored only if they have a minimum intron length of 30 nucleotides, have a minimum of 10 reads spanning each splice junction, above 10 uniquely mapping reads, and if they do not overlap with any exon; among other settings . SpliceQ @de_melo_costa_splice-q_2021 was used to quantify SE . The tool awards each intron an SE score ranging from zero to one. A score of zero means that the intron is fully retained, while a score of one means the complete splicing-out of the intron. Scores between zero and one indicate the approximate fraction of retained nucleotides.

## Data analyses

Analyses of the extracted SpliceQ data for all participants in the six studies were done using R version 4.4.1 on RStudio 2024.04.2+764 "Chocolate Cosmos" Release . Introns included in each data analyses step were only those whose SE was quantified across all samples.

## Investigating the impact of aging , and RT on splicing efficiency

`r length_perfect_scores` were excluded from the analyses as they had perfect scores across all samples.

To examine the effect of aging on splicing efficiency , we transformed participants' age to a scale of 0 to 1 and fitted a generalized linear mixed model (GLMM) using a Beta distribution. The model built using only the pre-exercise data included random intercepts for study and participants to account for random variations at the study and participant levels while sex and scaled age of participants were fixed effects. . Differentially spliced introns were defined as those with p values at or below 0.05.

To examine the effect of \~12 weeks RT on SE, we used a similar model that included fixed interactive effects for aging and RT, and sex. Study and participants remained random effects . Only samples that performed RT were included in this analyses. Thus, the control samples in ContraTrain, RELiEf and the non-RT samples in @robinson_enhanced_2017 were not included in the second model.

# References
