---
title: "Model_building"
format: html
editor: visual
---

## Introduction

This file contains the codes for the intron retention manuscript investigating the impact of aging and resistance training on intron retention in the skeletal muscle

```{r}
# echo: false
# Load the required packages
library(dplyr)
library(tidyverse)
library(seqwrap)
library(glmmTMB)



# Load the dataframe containing the 6 different RT intervention studies
# Check the /Data_extraction subfolder for how each individual study's data extraction was done

# Check "data_compilation.R for how data was compiled



all_splice_df <- readRDS("data/all_splice.RDS") %>%
  mutate(across(where(is.numeric), ~ round(.x, 2))) 



all_full_metadata <- readRDS("data/all_full_metadata.RDS")
colnames(all_full_metadata)


# REORDER THE SEQUENCE ID TO MATCH BOTH DATAFRAMMES
all_splice_reordered <- all_splice_df[, c("transcript_ID",all_full_metadata$seq_sample_id)] 

# Check if everything matches except the transcript_id
match(colnames(all_splice_reordered), all_full_metadata$seq_sample_id)



```

First, visualise the data

```{r}
library(ggplot2)

## Color scale ##
colors <-  c("#d7191c", "#fdae61", "#ffffbf", "#abd9e9", "#2c7bb6", "#1b9e77", "#d95f02", "#7570b3", "#e7298a", "#66a61e")


# plot a distribution of the participants in each study




# In the metadata, most participants contributed two samples

# subset the data to only count one participant once
meta_unique <- all_full_metadata %>%
  distinct(participant, .keep_all = TRUE)%>%
  mutate(study = recode(study,
                        "ReLiEf" = "RELIEF",
                        "copd" = "COPD",
                        "ct" = "ContraTRAIN",
                        "vol" = "VOLUME"))

# create counts per study
# Aim is to add number of participants  in image

counts_df <- meta_unique %>%
  group_by(study) %>%
  summarise(
    n = n(),
    male = sum(sex == "male"),
    female = sum(sex == "female"),
    .groups = "drop"
  )


# for the summarised image

sum_df <- meta_unique %>%
  summarise(
    n = n(),
    male = sum(sex == "male"),
    female = sum(sex == "female")
  )


# To help set in-image text
max_count <- ggplot_build(
  ggplot(meta_unique, aes(x = age)) +
    geom_histogram(binwidth = 5)
)$data[[1]]$count %>% max()




# Plot the distribution of all participants in one image
ggplot(meta_unique, aes(x = age, fill = sex)) +
  geom_histogram(position = position_dodge(width = 5),
                 alpha = 0.5,
                 binwidth = 5,
                 color = "black") +
  scale_fill_manual(
    values = c(
      "male" = colors[7],
      "female" = colors[3]
    )
  ) +
  geom_text(
    data = sum_df, aes(x = min(meta_unique$age), y= 0.5 * max_count,
                          label = paste0("n = ", n,
                                         "\nMales = ", male,
                                         "\nFemales = ", female)),
    inherit.aes = F,
    hjust = 0.5,
    vjust = 0.5,
    size = 4,
     fontface= "bold.italic"
  )+
 # facet_wrap(~ study, ncol = 2) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Age distribution of all participants",
    x = "Age",
    y = "Number of participants"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.title = element_blank()
  )



# ggsave("Figures/Dist_all_participants.png", bg = colors[4], scale = 2.5, dpi = 400)



ggplot(meta_unique, aes(x = age, fill = sex)) +
  geom_histogram(position = "dodge",
                 alpha = 0.5,
                 binwidth = 5,
                 color = "black") +
  scale_fill_manual(
    values = c(
      "male" = colors[7],
      "female" = colors[3]
    )
  ) +
  facet_wrap(~ study, ncol = 2, scales = "fixed") +
  geom_text(
    data = counts_df, aes(x = min(meta_unique$age), y= 0.5 * max_count,
                          label = paste0("n = ", n,
                                         "\nMales = ", male,
                                         "\nFemales = ", female)),
                          inherit.aes = F,
                          hjust = 0.1,
                          vjust = 1.5,
                          size = 4,
                         # fontface= "bold.italic"
  )+
  theme_minimal(base_size = 12) +
  labs(
    title = " Distribution of Participants across all studies",
    x = "Age",
    y = "Count"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.title = element_blank()
  )



# ggsave("Figures/Dist_all_participants_by_study.png", bg = colors[4], scale = 2.5, dpi = 400)
```

### Investigating a binary outcome

SPLICE-q scores the splicing efficiency at every junction 0-1 with one being a perfectly removed intron. This binary model investigates the question, "given an intron, what is the probability of perfect splicing as a function of age and resistance exercise training"

```{r}
#| echo: false

# derive a matrix that indicates 0 if SE is not 1
one_inflated_mat <- all_splice_reordered

one_inflated_mat[-1] <- lapply(
  one_inflated_mat[-1],
  function(x) as.integer(x == 1)
)




# Intialise argument
args_binom <- list( formula = y ~ scaled_age + time + sex +
                      (1 | study) + (1 | participant), family  = binomial)

# containerise using seqwrap_compose
binom <- seqwrap_compose(data       = one_inflated_mat,
                        metadata   = all_full_metadata,
                        samplename = "seq_sample_id",
                        modelfun   = glmmTMB::glmmTMB,
                        arguments  = args_binom)

# build model
binom_results <- seqwrap(binom,
                          return_models = FALSE,
                          cores = 10)

#saveRDS(binom_results, "data/binom_model.RDS")
# binom_results<- readRDS("data/binom_model.RDS")
```

### Second model

This model accepts as input the full spectrum of SE values. It investigates the impact of resistance training and aging on the slightest SE variations of introns.

```{r}
# convert the 1.0 to 0.999. This is becasue beta-model accepts only values between 0 and one
all_splice_reordered[all_splice_reordered == 1 ] <- 0.999



# initialise the argument. This time we check the interaction of age and time
args_full <-list(formula = y ~  scaled_age + time + sex + (1|study) +(1|participant), 
                 family = glmmTMB::beta_family(link = "logit"))




# check the functions and datasets
container <- seqwrap_compose(data = all_splice_reordered,
                             metadata = all_full_metadata,
                             samplename = "seq_sample_id",
                             modelfun = glmmTMB::glmmTMB,
                             arguments = args_full)


# build model
full_model <- seqwrap(container,
                 # summary_fun = sum_with_pred,
                 #eval_fun = eval_mod,
                 return_models = F,
                 # subset = 1:150,
                 cores = 10)


# full_model<- readRDS("data/full_model.RDS")

# saveRDS(full_model, "data/full_model.RDS")






```

### Visualising the results of both models

```{r}
#| warning: False  
library(dplyr)
library(tidyverse)
library(gridExtra)
library(ggpubr)
library(cowplot)
library(ggridges)
library(UpSetR)
library(ggplot2)
library(ggplotify)
library(clusterProfiler)
library(org.Hs.eg.db)
library(magick)




informed_binom_sum <- seqwrap_summarise(binom_results)


binom_model_outputs <- informed_binom_sum$summaries %>% 
  dplyr::select(-group) %>%
  dplyr::filter(term != "(Intercept)") %>%
  drop_na() %>%
  group_by(term) %>%                                  
  mutate(adj.p = p.adjust(p.value, method = "fdr")) %>% 
  ungroup() %>%
  # select significantly differentially spliced introns
  filter(adj.p <= 0.05) %>%
   mutate(term = recode(term,
                        "scaled_age" = "Aging",
                        "timePostExc" = "Resistance Exercise Training"))

# saveRDS(binom_model_outputs, "data/binom_model_outputs.RDS")

# subset those by age 

ds_introns_age <- binom_model_outputs %>%
  dplyr::filter(term == "Aging")

ds_introns_RT <- binom_model_outputs %>%
  dplyr::filter(term == "Resistance Exercise Training")




# extract the results of the non-binarised model

full_model_sum <- seqwrap_summarise(full_model)

# The non-binarised model 
full_model_outputs <- full_model_sum$summaries %>% 
  dplyr::select(-group) %>%
  dplyr::filter(term != "(Intercept)") %>%
  drop_na() %>%
  group_by(term) %>%                                  
  mutate(adj.p = p.adjust(p.value, method = "fdr")) %>% 
  ungroup() %>%
  # select significantly differentially spliced introns
  filter(adj.p <= 0.05) %>%
   mutate(term = recode(term,
                        "scaled_age" = "Aging",
                        "timePostExc" = "Resistance Exercise Training"))

#saveRDS(full_model_outputs, "data/full_model_outputs.RDS")
# x <- informed_binom_sum$summaries 
# length(unique(x$target))

full_ds_age <- full_model_outputs %>%
  dplyr::filter(term == "Aging")

full_ds_RT <- full_model_outputs %>%
  dplyr::filter(term == "Resistance Exercise Training")
```

```{r}
# identify the genes containing these introns and the intron lengths



# Load the gene annotation file
gene_annotation <- readRDS("data/ensembl_gene_annotation.RDS")

# Load one file from which we will extract intron length
# This is valid as only introns quantified in all samples were included in the analyses
intron_length <- readr::read_tsv("data_new/Alpha_Omega_SpliceQ_outputs/A_102.tsv") %>%
  
  distinct(across(6:ncol(.)), .keep_all = T) %>% # Removes duplicates based on columns 6 to end
  mutate(transcript_ID = paste0(transcript_ID, "_", intron_ID, "_", chr),
         intron_length = abs((sj3start - sj5end) + 1) ) %>% # Ensures positive length regardless of strand
  dplyr::select(transcript_ID, intron_length)


binom_model_outputs <- binom_model_outputs %>%
  inner_join(intron_length, by = c("target" = "transcript_ID")) %>%
  mutate(effect = case_when(estimate > 0 & adj.p <= 0.05 ~ "Improved SE", 
                            estimate < 0 & adj.p <= 0.05 ~ "Reduced SE" ,
                            estimate < 0 & adj.p > 0.05 ~ "No effect",
                            estimate > 0 & adj.p > 0.05 ~ "No effect")) %>%
  mutate(transcript_ID = str_split(target, "_",simplify= T) [,1]) %>%
  inner_join(gene_annotation, by= c("transcript_ID" = "ensembl_transcript_id_version"))


# plot distribution of phenotypes
binom_model_outputs %>%
  filter(term != "sexmale") %>%
  distinct(target, transcript_biotype, .keep_all = T) %>%
  ggplot(aes(transcript_biotype, fill = transcript_biotype))+
  geom_bar()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        plot.title = element_text(hjust = 0.5),
        legend.position = "none")+ # no legend
  stat_count(geom = "Text", aes(label = ..count..), vjust = -0.1) +
  ggtitle("Distribution of biotypes of genes containing the differentially spliced introns") +
  ylab("Number of introns per biotype") 
  


binom_model_outputs %>%
  filter(term != "sexmale") %>%
  dplyr::select(target, intron_length, effect) %>%
  ggplot(aes(x = intron_length, fill = effect)) +
  geom_histogram() +
  labs(
    title = "Distribution of Intron Length in model 1",
    x = "Intron Length",
    y = "Number of Introns"
  ) +
 # scale_x_continuous(limits = c(70, 30000)) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

min(binom_model_outputs$intron_length)

max(binom_model_outputs$intron_length)
# binom_model_outputs %>%
#   filter(term == "scaled_age") %>%
#   dplyr::select(target, intron_length, effect) %>%
#   ggplot(aes(x = intron_length, fill = effect)) +
#   geom_histogram() +
#   labs(
#     title = "Distribution of Intron Lengths in those affected by RT",
#     x = "Intron Length",
#     y = "Number of Introns"
#   ) +
#  # scale_x_continuous(limits = c(70, 30000)) +
#   theme_minimal() +
#   theme(plot.title = element_text(hjust = 0.5))




  
```

```{r}
# Evaluating the betabinomial model

full_model <- full_model_outputs %>%
  inner_join(intron_length, by = c("target" = "transcript_ID")) %>%
  mutate(effect = case_when(estimate > 0 & adj.p <= 0.05 ~ "Improved SE", 
                            estimate < 0 & adj.p <= 0.05 ~ "Reduced SE" ,
                            estimate < 0 & adj.p > 0.05 ~ "No effect",
                            estimate > 0 & adj.p > 0.05 ~ "No effect")) %>%
  mutate(transcript_ID = str_split(target, "_",simplify= T) [,1]) %>%
  inner_join(gene_annotation, by= c("transcript_ID" = "ensembl_transcript_id_version"))


# plot distribution of phenotypes
full_model %>%
  filter(term != "sexmale") %>%
  distinct(target, transcript_biotype, .keep_all = T) %>%
  ggplot(aes(transcript_biotype, fill = transcript_biotype))+
  geom_bar()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        plot.title = element_text(hjust = 0.5),
        legend.position = "none")+ # no legend
  stat_count(geom = "Text", aes(label = ..count..), vjust = -0.1) +
  ggtitle("Distribution of biotypes of genes containing the introns") +
  ylab("Number of introns per biotype") 
  


full_model %>%
  filter(term != "sexmale") %>%
  dplyr::select(target, intron_length, effect) %>%
  ggplot(aes(x = intron_length, fill = effect)) +
  geom_histogram() +
  labs(
    title = "Distribution of Intron Lengths in those affected by RT",
    x = "Intron Length",
    y = "Number of Introns"
  ) +
 # scale_x_continuous(limits = c(70, 30000)) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

# min(full_model$intron_length)
# max(full_model$intron_length)

full_model %>%
  filter(term == "Aging") %>%
  dplyr::select(target, intron_length, effect) %>%
  ggplot(aes(x = intron_length, fill = effect)) +
  geom_histogram() +
  labs(
    title = "Distribution of Intron Lengths in those affected by RT",
    x = "Intron Length",
    y = "Number of Introns"
  ) +
 # scale_x_continuous(limits = c(70, 30000)) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))



```

```{r}

df <- binom_model_outputs %>%
  dplyr::filter(term != "sexmale") %>%
  separate(target, into = c(NA, "intron_ID", NA), sep = "_") %>%
  mutate(
    gene_label = ifelse(
      is.na(external_gene_name) | external_gene_name == "",
      ensembl_gene_id,
      external_gene_name
    ),
    gene_intron = paste(gene_label, intron_ID, sep = " : ")
  )

ggplot(df, aes(x = estimate, y = reorder(gene_intron, estimate), color = effect)) +
  geom_point(size = 3) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  facet_wrap(~ term, scales = "free_y") +
  scale_color_manual(values = c("Improved SE" =  colors[2], "Reduced SE" = colors[1])) +
  labs(
    x = "Effect size",
    y = "Gene : Intron_ID",
    title = "Complete Intron Splicing Changes in Aging and Resistance Training",
   subtitle = "Binomial model (splicing efficiency coded as 0/1)"
  ) +
  theme_minimal() +
  theme(
  axis.text.y = element_text(size = 6),
  plot.title = element_text(hjust = 0.5),
  plot.subtitle = element_text(hjust = 0.5, size = 8)
)
ggsave("Figures/DS_introns_binary_model.png", bg = colors[4], scale = 2.5, dpi = 400)

```

```{r}
full <- full_model %>%
  dplyr::filter(term != "sexmale") %>%
  separate(target, into = c(NA, "intron_ID", NA), sep = "_") %>%
 # mutate(gene_intron = paste(external_gene_name, intron_ID, sep = " : "))
  mutate(
    gene_label = ifelse(
      is.na(external_gene_name) | external_gene_name == "",
      ensembl_gene_id,
      external_gene_name
    ),
    gene_intron = paste(gene_label, intron_ID, sep = " : ")
  )

ggplot(full, aes(x = estimate, y = reorder(gene_intron, estimate), color = effect)) +
  geom_point(size = 2) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  facet_wrap(~ term, scales = "free_y") +
  scale_color_manual(values = c("Improved SE" =  colors[2], "Reduced SE" = colors[1])) +
  labs(
    x = "Effect size",
    y = "Gene : Intron",
    title = "Intron Splicing Efficiency Changes in Aging and Resistance Training",
   # subtitle = "Only FDR-significant results shown"
  ) +
  theme_minimal()+
  theme(
    axis.text.y = element_text(size = 6),
    plot.title = element_text(hjust = 0.5)
  )

ggsave("Figures/DS_introns_beta_binomial_model.png", bg = colors[4], scale = 2.5, dpi = 400)


# library(dplyr)
# library(ggplot2)
# library(forcats)

df <- full %>%
  arrange(gene_label, estimate) %>%
  mutate(gene_intron = factor(gene_intron, levels = unique(gene_intron)))

ggplot(df, aes(x = estimate, y = gene_intron, color = effect)) +
  geom_point(size = 2) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  facet_wrap(~ term, scales = "free_y") +
  scale_color_manual(values = c("Improved SE" = colors[2],
                                "Reduced SE" = colors[1])) +
  labs(
    x = "Effect size",
    y = "Gene : Intron",
    title = "Intron-Specific Splicing Effects within Genes"
  ) +
  theme_minimal() +
  theme(
    axis.text.y = element_text(size = 6),
    plot.title = element_text(hjust = 0.5)
  )



# 
# ggplot(df, aes(x = estimate, y = gene_intron)) +
#   geom_point(aes(color = shared), size = 2) +
#   geom_vline(xintercept = 0, linetype = "dashed") +
#   facet_wrap(~ term, scales = "free_y") +
#   scale_color_manual(values = c("TRUE" = "red", "FALSE" = "grey70")) +
#   theme_minimal() +
#   labs(
#     x = "Effect size",
#     y = "Gene : Intron",
#     title = "Intron-Specific Splicing Effects within Genes",
#     subtitle = "Shared introns (affected by both aging and training) highlighted in red"
#   )
```

shared

```{r}
# plot those affected by both aging and exercise

shared <- full %>%
  group_by(gene_intron) %>%
  filter(n_distinct(term) == 2) %>%   # keeps only those seen in both terms
  ungroup()

shared$ensembl_gene_id
shared_gene <- full %>%
  group_by(gene_label) %>%
  filter(n_distinct(term) == 2) %>%
  ungroup() %>%
  arrange(gene_label, estimate) %>%   # gene first, then introns
  mutate(gene_intron = factor(gene_intron, levels = unique(gene_intron)))
  
# plot introns affected by both aging and RT
ggplot(shared, aes(x = estimate, y = reorder(gene_intron, estimate), color = effect)) +
  geom_point(size = 3) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  facet_wrap(~ term, scales = "free_y") +
  scale_color_manual(values = c("Improved SE" = colors[2],
                                "Reduced SE" = colors[1])) +
  labs(
    x = "Effect size",
    y = "Gene : Intron",
    title = "Introns Affected by Both Age and Training"
  ) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 8))



# plot genes affected by both aging and RT

ggplot(shared_gene, aes(x = estimate, y = gene_intron, color = effect)) +
  geom_point(size = 3) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  facet_wrap(~ term, scales = "free_y") +
  scale_color_manual(values = c("Improved SE" = colors[2],
                                "Reduced SE" = colors[1])) +
  labs(
    x = "Effect size",
    y = "Gene : Intron",
    title = "Introns Affected by Both Age and Training"
  ) +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 8))

ggsave("Figures/introns_affected_both_RT_and_aging.png", bg = colors[4], scale = 2.5, dpi = 400)


```

```{r}
# quantify the terms in each 
full %>%
  group_by(term, effect) %>%
  summarise(n = n_distinct(gene_label))


df %>%
  group_by(term, effect) %>%
  summarise(n = n_distinct(gene_label))


```

```{r}
ggplot(shared_gene, aes(x = term, y = estimate, group = gene_intron)) +
  geom_line(alpha = 0.4, color = "grey") +
  geom_point(aes(color = effect), size = 3) +
  scale_color_manual(values = c("Improved SE" = colors[2],
                                "Reduced SE" = colors[1])) +
  theme_minimal()


ggplot(full, aes(x = term, y = estimate, group = gene_intron)) +
  geom_line(alpha = 0.4, color = "grey") +
  geom_point(aes(color = effect), size = 3) +
  scale_color_manual(values = c("Improved SE" = colors[2],
                                "Reduced SE" = colors[1])) +
  theme_minimal() 





full %>%
  filter(term == "Aging" ) %>%
  arrange(desc(abs(estimate)))%>%
  select(gene_intron, effect,  estimate, gene_label )



full %>%
  filter(term == "Resistance Exercise Training") %>%
  arrange(desc(abs(estimate))) %>%
  select(gene_intron, effect,  estimate, gene_label )
```

\

```{r}
library(clusterProfiler)
library(org.Hs.eg.db)



# Functional annotation of genes affecetd by aging and RT

# First load the gene expression dataset
gene_exp_df <- readRDS("data_new/gene_counts/batch_corrected_genecounts.RDS") 

full_ds_RT <- full %>%
  filter(term == "Resistance Exercise Training")

# Functional annotation of the genes affected
ego_RT <- enrichGO(gene =  full_ds_RT$external_gene_name,
                      keyType = "SYMBOL",
                      universe = gene_exp_df$gene_name,
                      OrgDb = org.Hs.eg.db, 
                      ont = "BP", 
                      pAdjustMethod = "BH", 
                      qvalueCutoff = 0.05, 
                      readable = T)


## Output results from GO analysis to a table
cluster_RT <- data.frame(ego_RT)

go_RT <- dotplot(ego_RT,
                    
                    font.size = 8, title = "Enriched biological processes in genes containing introns with RT-associated SE") +
  theme(axis.text = element_text(size = 10), axis.text.y = element_text(size = 8), axis.title.x = element_text(size = 10),
        plot.title = element_text(hjust = 0) )

print(go_RT)

```

```{r}
full_ds_aging <- full %>%
  filter(term == "Aging")

# Functional annotation of the genes affected
ego_aging <- enrichGO(gene =  full_ds_aging$external_gene_name,
                      keyType = "SYMBOL",
                      universe = gene_exp_df$gene_name,
                      OrgDb = org.Hs.eg.db, 
                      ont = "BP", 
                      pAdjustMethod = "BH", 
                      qvalueCutoff = 0.05, 
                      readable = T)


## Output results from GO analysis to a table
cluster_aging <- data.frame(ego_aging)

go_aging <- dotplot(ego_aging,
                    
                    font.size = 8, title = "Enriched biological processes in genes containing introns with aging SE") +
  theme(axis.text = element_text(size = 10), axis.text.y = element_text(size = 8), axis.title.x = element_text(size = 10),
        plot.title = element_text(hjust = 0) )

print(go_aging)

```

```{r}
age_introns <- full_model_outputs %>%
  dplyr::filter(term == "Aging") %>%
  dplyr::arrange(desc(abs(estimate)))%>%
  dplyr::slice(1:10) # %>%
 # pull(target)
  


age_introns_df <- all_splice_df %>%
  dplyr::filter(transcript_ID %in% age_introns$target) %>%
    pivot_longer(names_to = "seq_sample_id",
               values_to = "SE",
               cols = -(transcript_ID) ) %>%
  inner_join(age_introns, by = c("transcript_ID" = "target")) %>%
  separate(transcript_ID, into = c("transcript_ID", "intron_ID", NA), sep = "_") %>%
  inner_join(gene_annotation, by= c("transcript_ID" = "ensembl_transcript_id_version")) %>%
   
  mutate(
    gene_label = ifelse(
      is.na(external_gene_name) | external_gene_name == "",
      ensembl_gene_id,
      external_gene_name
    ),
    gene_intron = paste(gene_label, intron_ID, sep = " : ")) %>%
  inner_join(all_full_metadata, by = "seq_sample_id")%>%
  group_by(scaled_age, gene_intron) %>%
  summarise(mean_SE = mean(SE, na.rm = TRUE), .groups = "drop")

ggplot(age_introns_df, aes(x = scaled_age, y = mean_SE, color = gene_intron)) +
  geom_point(alpha = 0.4, size = 1) +   # ???? dots instead of lines
  geom_smooth(aes(group = 1), method = "lm", color = "red", size = 1.2) +
  theme_minimal()


ggplot(age_introns_df, aes(x = scaled_age, y = mean_SE, group = gene_intron)) +
  geom_point(alpha = 0.3, size = 0.8) +
  geom_smooth(aes(color = gene_intron), method = "lm", se = FALSE, size = 0.8) +
  theme_minimal() +
  theme(legend.position = "none")

```

```{r}
RT_introns <- full_model_outputs %>%
  dplyr::filter(term == "Resistance Exercise Training") %>%
  dplyr::arrange(desc(abs(estimate)))%>%
  dplyr::slice(1:10) # %>%
 # pull(target)
  


RT_introns_df <- all_splice_df %>%
  dplyr::filter(transcript_ID %in% RT_introns$target) %>%
    pivot_longer(names_to = "seq_sample_id",
               values_to = "SE",
               cols = -(transcript_ID) ) %>%
  inner_join(RT_introns, by = c("transcript_ID" = "target")) %>%
  separate(transcript_ID, into = c("transcript_ID", "intron_ID", NA), sep = "_") %>%
  inner_join(gene_annotation, by= c("transcript_ID" = "ensembl_transcript_id_version")) %>%
   
  mutate(
    gene_label = ifelse(
      is.na(external_gene_name) | external_gene_name == "",
      ensembl_gene_id,
      external_gene_name
    ),
    gene_intron = paste(gene_label, intron_ID, sep = " : ")) %>%
  inner_join(all_full_metadata, by = "seq_sample_id") %>%
   group_by(time, gene_intron) %>%
   summarise(mean_SE = mean(SE, na.rm = TRUE), .groups = "drop")


ggplot(RT_introns_df, aes(x = time, y = mean_SE, colour = gene_intron)) +
  geom_point(outlier.shape = NA) +
  geom_jitter(alpha = 0.2, width = 0.5, size = 1) +
  theme_minimal()



```

```{r}
# explore how the aging-affected genes were expressed in relation to aging

gene_exp_df

x <- full_ds_aging %>%
   dplyr::arrange(desc(abs(estimate)))%>%
  dplyr::slice(1:5) 

exp_df <- gene_exp_df %>%
  dplyr::filter(gene_name %in% x$gene_label) %>%
    pivot_longer(names_to = "seq_sample_id",
               values_to = "gene_count",
               cols = -(gene_name) ) %>%
  inner_join(all_full_metadata, by = "seq_sample_id") %>%
  group_by(scaled_age, gene_name) %>%
  summarise(mean_count = mean(gene_count, na.rm = TRUE), .groups = "drop")




ggplot(exp_df, aes(x = scaled_age, y = mean_count, colour= gene_name)) +
  geom_point(outlier.shape = NA) +
  geom_jitter(alpha = 0.2, width = 0.5, size = 1) +
  theme_minimal()

```
