---
title: "Impact of age and resistance exercise training on intron retention"
format:
  docx:
   reference-doc: resources/custom_reference.docx
author:
  - "Chidimma Phoebe Echebiri^a,b^, Kristian Lian^b^, Torkil Rognefl&aring;ten^b^, 
 H&aring;vard Hamarsland^b^, Knut Sindre M&#248;lmen^b^, Rafi Ahmad^a^, Stian Ellefsen^b^, and Daniel Hammarstr&ouml;m^b^"
editor: visual
csl: resources/journal-of-cell-science.csl
bibliography: resources/My_Library.bib
target_journal: https://journals.biologists.com/jcs
prefer-html: true
---

::: {custom-style="author-meta"}
^a^ Department of Biotechnology, University of Inland Norway, Faculty of Applied Ecology, Agricultural Sciences and Biotechnology, Hamar, Norway

^b^ Section for Health and Exercise Physiology, Department of Public Health and Sport Sciences, University of Inland Norway, Lillehammer, Norway

Correspondance: Chidimma Phoebe Echebiri (chidimma.echebiri\@inn.no), Daniel Hammarström (daniel.hammarstrom\@inn.no).
:::

```{r}
#| label: setup
#| include: false


library(dplyr)
library(tidyverse)
library(gt)
library(cowplot)
library(ggpubr)
library(ggplot2)
library(gridExtra)
library(knitr)

color_scale <- c("#7fc97f",  "#beaed4","#fdc086","#ffff99","#386cb0","#f0027f")



source("R/Trainome_functions.R")

knitr::opts_chunk$set(warning = FALSE, message = FALSE)

```

::: {custom-style="abstracttitle"}
Abstract
:::

<div>

</div>

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: abstract

#source("R/Data_exploration.R")
  
  pre_metadata <- readRDS("data_new/Pre_Exercise/all_prexercise_metadata.RDS")
  
  participants <- length(unique(pre_metadata$participant))
  
  desc_stat  <- pre_metadata %>%
    distinct(sex, age, participant) %>%
  
    summarise(.by = c(sex),
              n= n(),
              range = range(age)) %>%
    data.frame()
  
  females <- glue::glue(round(desc_stat[1,2], 0),
                        " females aged ", round(desc_stat[1,3], 0),
                        " to ", round(desc_stat[2, 3], 0))
  
  males <- glue::glue( round(desc_stat[3,2], 0),
                        " males aged ", round(desc_stat[3,3], 0),
                      " to ", round(desc_stat[4, 3], 0))
  
  
  
  full_metadata <- readRDS("data_new/processed_data/all_full_metadata.RDS")
  full_part <- length(unique(full_metadata$participant))
  
  desc_stat_full  <- full_metadata %>%
    distinct(sex, age, participant) %>%
  
    summarise(.by = c(sex),
              n= n(),
              range = range(age)) %>%
    data.frame()
  
  females_full <- glue::glue(round(desc_stat_full[1,2], 0),
                        " females aged ", round(desc_stat_full[1,3], 0),
                        " to ", round(desc_stat_full[2, 3], 0))
  
  males_full <- glue::glue( round(desc_stat_full[3,2], 0),
                        " males aged ", round(desc_stat_full[3,3], 0),
                      " to ", round(desc_stat_full[4, 3], 0))
  
```

Progressive resistance exercise training (RT) elicits a range of adaptive responses in the human transcriptome, most of which have been studied in-depth. Several studies have shown that processes such as aging, disease conditions and physical exercise affect the pre-mRNA splicing machinery. There is however limited information about intron retention in healthy human skeletal muscle and how it is impacted by aging and resistance exercise training. This study investigates intron retention in the *vastus lateralis* muscle of `r participants` participants ( `r females` ; `r males`) ; and explores if and how resistance exercise training affects splicing efficiency in the `r  full_part` participants (( `r females_full` ; `r males_full`) who completed \~ 12 weeks of training . Using SpliceQ, splicing efficiency of introns in paired_end RNA sequence data from six different RT intevrntion studies involving 367 partricipants. Differentially spliced introns at baseline and at post exercise were identified using betabinomial generalised linear models allowing for study and participant level intercepts. The differentially spliced introns were further investigated to identify patterns related to their gene lengths, ontology and functions.

## Introduction

Aging is characterized by a progressive decline in most biological functions leading to increasing morbidity and mortality [@lopez-otin_hallmarks_2013; @keshavarz_targeting_2023 ; @gyenis_genome-wide_2023; @schaum_ageing_2020] . Physiologically, aging is associated with the induction of different stress response pathways in different cells, tissues and species [@stegeman_transcriptional_2017; @welle_gene_2003]. Aging and aging-like phenotypes are also associated with increased mis-splicing [@angarola_splicing_2021 ; @garcia-ruiz_splicing_2023 ; @nishida_tissue-_2015 ; @mariotti_deterioration_2022; @stegeman_transcriptional_2017; @pabis_concerted_2024; @latorre_splicing_2017; @holly_changes_2013] as a result of alterations in the expression of splicing factors ; a resilience strategy aimed at counteracting aging-induced damages and loss of function [@ferrucci_energysplicing_2022 ; @ubaida-mohien_discovery_2019].

Intron retention (IR) is the major form of alternative splicing that is affected by aging [@kodama_systematic_2024; @mariotti_deterioration_2022; @bhadra_alternative_2020]. It mostly affects protein-coding genes thus resulting in a transcriptomic and proteomic diversity that could influence cellular plasticity. IR occurs when a full intron, or parts of an intron in a mature messenger RNA (mRNA) is retained during splicing, resulting in the presence of an unprocessed sequence in the mRNA. The retention leads either to the degradation of the IR-containing transcripts through nonsense-mediated decay or the action of nucleases; or to frameshifts that cause the generation of alternative protein isoforms [@baralle_age-related_2023 ; @zheng_intron_2020 ; @mauger_targeted_2016; @wong_intron_2022 ; @monteuuis_changing_2019; @jacob_intron_2017 ]. IR predominantly occurs in transcripts that lack physiological relevance in cells and tissues they are detected in [@braunschweig_widespread_2014]. Increase in IR is a signature of the aging process [@bhadra_alternative_2020; @adusumalli_increased_2019; @baralle_age-related_2023] and is often measured in terms of splicing efficiency which describes the amount of introns retained in a mature mRNA. Splicing efficiency is cell and tissue specific [@nishida_tissue-_2015; @baralle_age-related_2023] but tends to be similar between the introns of a gene [@sanchez-escabias_co-transcriptional_2022] . Besides aging, poor splicing efficiency has also been linked to human diseases like alpha thalassemia [@nelson_novel_2005] , inflammatory bowel disease [@hasler_alterations_2011] , neurodegenerative diseases [@sznajder_intron_2018; @ong_increased_2020 ] and an increased risk of certain cancers [@dvinge_widespread_2015 ; @inoue_minor_2021 ; @shah_towards_2022].

Progressive resistance exercise training (RT) describes the type of physical exercise training where the skeletal muscle is exercised against progressively increased types of resistance [@liu_progressive_2009]. The benefits of RT include improvements in muscle strength, muscle mass, bone density, mRNA splicing [@kodama_systematic_2024] and several other major hallmarks of aging [@garatachea_exercise_2015 ]. RT could thus be a potent preventive or treatment strategy for improving aging-related functional disabilities associated with the aging skeletal muscle [@peterson_influence_2011 ; @kryger_resistance_2007 ; @lu_effects_2021 ]as well as for improving cardiorespiratory fitness in older persons [@smart_role_2022]. A lot is currently known about adaptive responses to RT. However, intron retention in response to RT is relatively unexplored in healthy individuals . If increased intron retention is linked to disease conditions and RT correlates to improved health condition, it should be important to investigate if the benefits of RT include improved splicing efficiency. This study is aimed at investigating differentially spliced introns in human *vastus lateralis* with increasing age, and how about 12 weeks of RT can affect splicing efficiency. This study will hopefully generate more insights into the molecular mediators of response to RT.

# Results

```{r}
#| echo: false
#| message: false
#| warning: false
#| label: Results

# Load the poorly spliced introns table
low_SE <- readRDS("data_new/Pre_Exercise/annotated_low_SE_introns.RDS") %>%
  dplyr::select(transcript_ID, intron_ID, transcript_biotype,external_gene_name, mode, min, max)
kable(low_SE, caption = " Table 1:Introns with poor splicing efficiency across all baseline samples")


perfect_SE <- readRDS("data_new/Pre_Exercise/annotated_perfect_SE_introns.RDS")

PE_introns <- length(perfect_SE$intron_ID)

PE_transcripts <- length(unique(perfect_SE$transcript_ID))

PE_genes <- length(unique(perfect_SE$ensembl_gene_id))
```

#### Some introns have characteristically poor or perfect splicing efficiency

To determine if there were introns having same splicing efficiency patterns across all baseline samples, we investigated those with poor splicing efficiency, as well as those with perfect splicing efficiency across all samples. To identify introns with characteristically low splicing efficiency , we filtered the baseline dataset for introns whose splicing efficiency across all baseline samples have a maximum total splicing efficiency of 0.2 . Six introns belonging to six different transcripts were shown to be almost 100% retained in all samples. The introns are as shown in **Table 1** **.** 5 of the 6 transcripts containing the poorly spliced introns are long non-coding RNAs without functional annotation on the org.Hs.eg.db database @manual_orghsegdb_nodate . The other transcript containing the poorly spliced intron is an alternatively spliced transcript of the MEIOC gene that lacks a coding sequence (CDS). GO analysis of the MEIOC transcript reveals it is involved in meiotic processes as seen in **Figure 1**. Only two of the poorly spliced introns were captured in the post exercise data where they still remained poorly spliced.

`r PE_introns` introns from `r PE_transcripts` transcripts belonging to `r PE_genes` genes were perfectly spliced across all datasets. These genes were mostly protein_coding genes enriched for biological processes that are physiologically relevent to the skeletal muscle as can be seen in **Figure 2**. 227 of these introns were captured in the postexercise data

In the postexercise data, 329 of the 8738 introns were perfectly spliced across all introns , 218 of which were among those perfectly spliced at baseline.

### Transcript length did not correlate to splicing efficiency

To investigate if there was a correlation between transcript length and splicing efficiency at baseline, a pearson's product-moment correlation test was performed and a negligible correllation of -0.018 at a pvalue of 0.007 was found between the transcript length and splicing efficiency. When tested in the post exercise data alone , it had a correlation estimate of -0.054 at a p value of 0.0. With the full dataset (baseline and postexercise), the correlation was -0.042 with a p.value of 0.0

When both post and preexercise data were included, 8738 introns were captured across all datasets

We measured intron retention as splicing efficiency using SpliceQ @demelocosta2021. mis-splicing occurs more in non-protein coding genes than in protein-coding genes @garcia-ruiz_splicing_2023, this disagrees with the results we obtained . We however take cognisance of the limitations in RNA-Seq's reported ability to correctly quantify non-coding RNAs @stokes_transcriptomics_2023.

### Differentially spliced introns at baseline

of the 23063 introns captured across all samples at baseline, 1158 were differentially spliced based on age. Gene ontology analyses reveal that the introns that tend to get poorly spliced with age tend to be related to skeletal muscle processes.

### Resistance Exercise training elicits a different set of ds intron

942 introns belonging to 580 transcripts belonging to 97 genes. were differentially spliced from the impact of RT, either alone (550) or in interaction with age (392)

## Materials and methods

#### Datasets

Six different studies designed to investigate different aspects of adaptive response to RT were included in the analyses. Five of these studies were conducted by our research group, while one is a study by @robinson_enhanced_2017 with GEO ascension number GSE97084. The studies were selected based on the criteria that they included paired-end Illumina RNA-Seq data from the *vastus lateralis* muscle of human subjects who performed at least 10 weeks of RT with samples collected at baseline and at the end of the study . Age of all participants was also required to be included in the data. The study protocols of the different studies are described elsewhere.

The characteristics of included studies are presented in **Table X** below

**Table 3**: basic characteristics of RT intervention studies included in this analysis: primary results from three of the RT intervention studies (COPD, Volume; SRP102542) have been previously published. Results from ContraTrain, Alpha & Omega, and ReLiEf are yet to be published but links to description of their study protocols are provided

+--------------+--------------------------+-------------------------------+------------------------+------------------------------------+
| Study        | Participants at baseline | participants at post-exercise | Age range              | Publication or clinical identifier |
+==============+==========================+===============================+========================+====================================+
| COPD         | 53                       | 53                            | 56-79                  | @molmen_chronic_2021               |
+--------------+--------------------------+-------------------------------+------------------------+------------------------------------+
| Volume       | 25                       | 25                            | 20-37                  | @khan_increased_2020               |
+--------------+--------------------------+-------------------------------+------------------------+------------------------------------+
| Contratrain  | 32                       | 26                            | 19-35                  | NCT03795025                        |
+--------------+--------------------------+-------------------------------+------------------------+------------------------------------+
| Alpha/ Omega | 143                      | 143                           | 23-60                  | NCT04279951                        |
+--------------+--------------------------+-------------------------------+------------------------+------------------------------------+
| SRP102542    | 52                       | 52                            | 19-30 (Young cohort)   | @robinson_enhanced_2017            |
|              |                          |                               |                        |                                    |
| (GSE97084)   |                          |                               | 65-78 (Old cohort)     |                                    |
+--------------+--------------------------+-------------------------------+------------------------+------------------------------------+
| ReLiEf       | 73                       | 66                            | 20 - 29 (Young cohort) | @lian_relief_nodate                |
|              |                          |                               |                        |                                    |
|              |                          |                               | 72 - 93 (Old cohort)   |                                    |
+--------------+--------------------------+-------------------------------+------------------------+------------------------------------+

### Methodology

#### RNA sequence data processing

The quality of the reads were checked using FASTQC(v0.11.9). STAR (2.7.9a) [@dobin2013] was used to map the RNA Sequence reads to the human reference genome (GRCh38, gencode v47 primary annotaion ) using the 2-pass alignment method described by @veeneman_two-pass_2016. This method is said to advance the quantification and discovery of splicing events. Expression levels of both genes and transcripts were quantified using RSEM (1.3.3) [@li_rsem_2011]. Splicing efficiency of each intron was calculated using the default settings in SpliceQ @de_melo_costa_splice-q_2021 . SpliceQ gives each intron an SE score ranging from zero to one. An SE of 0 indicates that the intron is fully retained, while 1 indicates the complete splicing-out of the intron. SE values inbetween 0 and 1 indicate the approximate fraction of nucleotides of an intron that was retained .

## Data analyses

Analyses of the extracted SpliceQ and RSEM data for all participants in the six studies were done using R version 4.4.1 on RStudio 2024.04.2+764 "Chocolate Cosmos" Release . Introns included in each data analyses step were only those whose splicing efficiency was captured across all samples.

#### Investigating the impact of aging , and RT on splicing efficiency

To examine the effect of aging on splicing efficiency, we transformed participants' age to a scale of 0 to 1 and fitted a generalized linear mixed model using a Beta distribution. The model included random intercepts for study and participants to account for random variations at the study and participant levels. The model also included random slopes for age within each study . The differentially spliced introns were defined as those with p values at or below 0.05 and log2Fc above 0.5. This was done using only baseline (pre-exercise) data.

The examine the effect of \~12 weeks RT on SE, we fitted a GLMM to examine the effect of RT. The model included fixed interactive effects for aging and RT, and random intercepts for study and participant, as well as random slopes for age within each study. Only samples that performed RT were included in post-exercise data were included in this analyses. Thus, the control samples in ContraTrain and the non-RT samples in @robinson_enhanced_2017 were not included in the second model.

As recommended by [@zhao2020] in scenerios where multiple datasets are combined, we used the class-specific strategy of quantile normalization to normalize the gene expression data.

## Discussion

splicing efficiency improves with increased expression level of genes [@saudemont_fitness_2017] and with longer genes which tend to agree that intron retention in normal circumstances are part of posttranscriptional regulation. IR reduces the expression of physiologically less relevant genes in a particular cell or tissue @braunschweig_widespread_2014 . Poor splicing efficiency of physiologically relevant genes can be deleterious thus its prevalence in less relevant genes are considered a mechanism with which gene expression is regulated. Aging is associated with the induction of different stress response pathways in different cells, tissues and species [@stegeman_transcriptional_2017; @welle_gene_2003]. IR is a physiological mechanism that controls gene expression [@wong_intron_2016 ; @schmitz_intron_2017 ]

Physical exercise helps maintain muscle health, general well-being. and has been linked to a reduced risk of with a reduced risk of all-cause mortality and cardiovascular diseases @ekelund_dose-response_2019 .

Thus, it plays roles in gene regulation, normal physiology and disease conditions [@wong_intron_2016; @ge_functional_2014; @inoue_minor_2021].

Splicing efficiency is cell and tissue specific [@nishida_tissue-_2015; @baralle_age-related_2023] but tends to be similar between the introns of a gene [@sanchez-escabias_co-transcriptional_2022] . This suggests that a balanced study of intron retention should compare same introns in same genes and same tissues across the subjects of interest.

IR is also linked to the inactivation of tumour supressor genes [@jung_intron_2015]

IR affects about 80% of human protein-coding genes [@middleton_irfinder_2017 ; @braunschweig_widespread_2014 ] and . High IR levels reportedly correlate with the down-regulation of mRNA transcripts in myelocytes and granulocytes @wong_orchestrated_2013 . Short genes with high mRNA levels tend to be inefficiently spliced compared to long genes with high mRNA levels . That is, there is a positive correlation between gene length, expression level and splicing efficiency (SE) [@sanchez-escabias_co-transcriptional_2022 , @garcia-ruiz_splicing_2023, @saudemont_fitness_2017, @braunschweig_widespread_2014] .

IR tends to increasingly occur in response to specific developmental cues and serve as a strategy for rapid mobilization of some mRNAs for protein translation [@ong_increased_2020].

This data suggests that SE is more efficient

Alternative splicing (AS) ; a mechanism in mRNA processing in which protein-coding and non-protein coding parts of the RNA are alternatively skipped results to a transcriptomic and proteomic diversity that determines cellular plasticity [@holly_changes_2013; @angarola_splicing_2021; @schwerk_regulation_2005 .]

Exercise counteracts the effects of aging on mitochondrial health and can offset aging-related changes to the splicing machinery [@ubaida-mohien_physical_2019; @ubaida-mohien_unbiased_2022]

Aging is often assessed using the concept of a "biological clock" that is organ, system and sex specific [@nie_distinct_2022].

IR is a physiological mechanism that controls gene expression [@wong_intron_2016 ; @schmitz_intron_2017 ] increase in IR is a signature of the aging process @bhadra_alternative_2020

Tissues and organs are altered differently by aging and aging-related processes @baralle_age-related_2023 .

[@sanchez-escabias_co-transcriptional_2022 ; @garcia-ruiz_splicing_2023, @saudemont_fitness_2017] suggest a correlation between physiologic relevance, gene length and splicing efficiency. The generated data agrees with that as the introns poorly spliced across all samples were associated with biological processes of less relevance to the skeletal muscle when compared to those in the perfectly spliced introns. The gene ontology of the genes contaning the poorly spliced introns, and those containing the perfectly spliced introns

@saudemont_fitness_2017 stipulated a mutation-selection-drift theory that suggests shorter, intron-poor and lowly expressed genes show a poorer splicing efficiency

### Shortcoming

This study did not group RT into conditions of volume of exercise done. The benefits of RT are known to be volume-dependent and the 6 studies involved varying RT conditions
