---
title: "Data analysis:"
format: html
bibliography: "My_Library.bib"
execute: 
  warning: false
  message: false
  code-summary: "Code"
  code-fold: true
editor_options: 
  chunk_output_type: console
---

```{r}
#| code-fold: true
#| code-summary: "Preliminary settings and packages"


color_scale <- c("#7fc97f",  "#beaed4","#fdc086","#ffff99","#386cb0","#f0027f")

library(dplyr)
library(gt)
library(tidyverse)
library(cowplot)
library(seqwrap)


source("R/Trainome_functions.R")


```

## Introduction

This file reproduces the relevant analyses for the paper as well as the supplementary materials

### Data Extraction

Running the codes in this section extracts and processes the data for the individual datasets. It also saves the extracted datasets in the designated folders for further analyses

```{r}
# Extract the Alpha and Omega dataset
source("R/Data_extraction/Alpha_Omega_data.R")

# Extract the COPD dataset
source("R/Data_extraction/COPD_data.R")

```

The `echo: false` option disables the printing of code (only output is displayed).
