---
title: "Data analysis:"
format: html
bibliography: "My_Library.bib"
execute: 
  warning: false
  message: false
  code-summary: "Code"
  code-fold: true
editor_options: 
  chunk_output_type: console
---

```{r}
#| code-fold: true
#| code-summary: "Preliminary settings and packages"


color_scale <- c("#7fc97f",  "#beaed4","#fdc086","#ffff99","#386cb0","#f0027f")

library(dplyr)
library(gt)
library(tidyverse)
library(cowplot)


source("R/Trainome_functions.R")


```

## Introduction

This file reproduces the relevant analyses for the paper as well as the supplementary materials

### Data Extraction

Running the codes in this section extracts and processes the data for the individual datasets. It also saves the extracted datasets in the designated folders for further analyses

```{r}
#Create these directories if the don't already exist

# A folder for baseline data
ifelse(!dir.exists(file.path("data_new", "Pre_Exercise")), dir.create(file.path("data_new", "Pre_Exercise")), FALSE)

# A folder for the full datasets
ifelse(!dir.exists(file.path("data_new", "processed_data")), dir.create(file.path("data_new", "processed_data")), FALSE)


# Extract the Alpha and Omega data
source("R/Data_extraction/Alpha_Omega_data.R")

# Extract the COPD data
source("R/Data_extraction/COPD_data.R")

# Extract the Contratrain data
source("R/Data_extraction/ct_data.R")

# Extract the Volume data
source("R/Data_extraction/Volume_data.R")


# Extract the Public data; Srp102542
source("R/Data_extraction/SRP102542_data.R")

```

Running the function above would save the following results in a `data_new` folder

| Dataset                                                      | Description                       |
|----------------------------------------------|--------------------------|
| `data_new/processed_data/Alpha_Omega_splicing_data.RDS`      | Alpha_Omega SpliceQ data          |
| `data_new/processed_data/Alpha_Omega_metadata.RDS`           | Alpha_Omega metadata              |
| `data_new/Pre_Exercise/Alpha_Omega_PreExc_metadata.RDS`      | Alpha_Omega baseline metadata     |
| `data_new/Pre_Exercise/Alpha_Omega_PreExc_splicing_data.RDS` | Alpha_Omega baseline SpliceQ data |
| `data_new/processed_data/copd_metadata.RDS`                  | COPD metadata                     |
| `data_new/processed_data/copd_splicing_data.RDS`             | COPD SpliceQ data                 |
| `data_new/Pre_Exercise/copd_preExc_metadata.RDS`             | COPD baseline metadata            |
| `data_new/Pre_Exercise/copd_preExc_splicing_data.RDS`        | COPD baseline SpliceQ data        |
| `data_new/processed_data/contratrain_metadata.RDS`           | Contratrain metadata              |
| `data_new/Pre_Exercise/ct_PreExc_metadata.RDS`               | Contratrain baseline metadata     |
| `data_new/processed_data/contratrain_splicing_data.RDS`      | Contratrain SpliceQ data          |
| `data_new/Pre_Exercise/ct_PreExc_splicing_data.RDS`          | Contratrain baseline SpliceQ data |
| `data_new/processed_data/volume_metadata.RDS`                | Volume metadata                   |
| `data_new/processed_data/volume_splicing_data.RDS`           | Volume SpliceQ data               |
| `data_new/Pre_Exercise/vol_preExc_metadata.RDS`              | Volume baseline metadata          |
| `data_new/Pre_Exercise/vol_preExc_splicing_data.RDS`         | Volume baseline SpliceQ data      |
| `data_new/Pre_Exercise/SRP102542_preExc_metadata.RDS`        | SRP102542 baseline metadata       |
| `data_new/processed_data/SRP102542_metadata.RDS`             | SRP102542 metadata                |
| `data_new/processed_data/SRP102542_splicing_data.RDS`        | SRP102542 SpliceQ data            |
| `data_new/Pre_Exercise/SRP102542_preExc_splicing_data.RDS`   | SRP102542 baseline SpliceQ data   |
