---
title: "Data analysis:"
format: html
execute: 
  warning: false
  message: false
  code-summary: "Code"
  code-fold: true
editor_options: 
  chunk_output_type: console
---

```{r}
#| code-fold: true
#| code-summary: "Preliminary settings and packages"


color_scale <- c("#7fc97f",  "#beaed4","#fdc086","#ffff99","#386cb0","#f0027f")

library(dplyr)
library(gt)
library(tidyverse)
library(cowplot)


source("R/Trainome_functions.R")


```

## Introduction

This file reproduces the relevant analyses for the paper as well as the supplementary materials

### Data Extraction

Running the codes in this section extracts and processes the data for the individual datasets. It also saves the extracted datasets in the designated folders for further analyses

```{r}
#Create these directories if the don't already exist


# Create data_new directory if it doesnt exist
if (!dir.exists("data")) {
  dir.create("data")
}


# Extract the Alpha and Omega data
source("R/Data_extraction/Alpha_Omega_data.R")

# Extract the COPD data
source("R/Data_extraction/COPD_data.R")

# Extract the Contratrain data
source("R/Data_extraction/ct_data.R")

# Extract the Volume data
source("R/Data_extraction/Volume_data.R")


# Extract the Public data; Srp102542
source("R/Data_extraction/SRP102542_data.R")


# Extract the REliEf data
source("R/Data_extraction/Relief.R")
```

Running the function above would save the following results in a `data_new` folder

| Data                                 | Description              |
|--------------------------------------|--------------------------|
| `data/Alpha_Omega_splicing_data.RDS` | Alpha_Omega SpliceQ data |
| `data/Alpha_Omega_metadata.RDS`      | Alpha_Omega metadata     |
| `data/copd_metadata.RDS`             | COPD metadata            |
| `data/copd_splicing_data.RDS`        | COPD SpliceQ data        |
| `data/contratrain_metadata.RDS`      | Contratrain metadata     |
| `data/contratrain_splicing_data.RDS` | Contratrain SpliceQ data |
| `data/volume_metadata.RDS`           | Volume metadata          |
| `data/volume_splicing_data.RDS`      | Volume SpliceQ data      |
| `data/SRP102542_metadata.RDS`        | SRP102542 metadata       |
| `data/SRP102542_splicing_data.RDS`   | SRP102542 SpliceQ data   |
| `data/Relief_splicing_data.RDS`      | ReLiEf SpliceQ data      |
| `data/Relief_metadata.RDS`           | ReLiEf metadata          |

### Basic data exploration

Running this code would perform basic data exploration using the data extracted and saved above. It also saves a couple of images and files as listed in the table below the code.

```{r}
# 
source("R/new_data_extraction.R")
```

The code block above would produce and save the following files and images

| Data                               | Description                                     |
|------------------------------------|-------------------------------------------------|
| `data/all_full_metadata.RDS`       | Combined metadata of all the datasets           |
| `data/all_splice.RDS`              | All splicing data in one dataframe              |
| `data/ensembl_gene_annotation.RDS` | All genes annotations (downloaded from Ensembl) |

### Model building

This script in the code block would build a betabinomial GLMM model to analyse differentially spliced introns based on age (scaled value), resistance exercise training and its interaction with aging while having participant and study as random factors

```{r}


source("R/model_building.R")
```

The script above produces one file "data/RT_model_summary.RDS" that contains the model summary and predictions

### Model Exploration

The following script would explore the model outputs and create the figures and files shown in the table below

```{r}
source("R/model_exploration.R")
```

| Data                         | Argument                        |
|------------------------------|---------------------------------|
| `"data_new/RT_model_df.RDS"` | The model and model predictions |
|                              |                                 |
|                              |                                 |
|                              |                                 |
|                              |                                 |
|                              |                                 |
|                              |                                 |
