---
title: "Visualisation"
format: pdf
editor: source
---

## Introduction

Using SPLICE-q, intron splicing efficiency was quantified from RNA Sequence data generated from five different resistance exercise training (RT) intervention studies . The study was aimed at investigating if and how aging, and, RT affected intron splicing efficiency in human skeletal muscle. Due to an inability to separate study effects from the effects of aging in the pooled dataset as participants in majority of the studies had distinct age distributions, this study employed a two-step complementary approach; a primary analysis using the RELIEF data which has young and old participants, and a secondary analysis using all 5 studies where the trajectory of differentially spliced introns from the primary analysis will be mapped.

### Basic data visualization

#### Figure 2

```{r}
#| echo: false
# load the file that contains all the data
source(here::here("R/Load_data_for_visualisation.R"))

# plot a distribution of the participants in each study

# In the metadata, most participants contributed two samples

# subset the data to only count one participant once
meta_unique <- metadata %>%
  distinct(participant, .keep_all = TRUE)%>%
  mutate(study = recode(study,
                        "ReLiEf" = "RELIEF",
                        "copd" = "COPD",
                        "ct" = "ContraTRAIN",
                        "vol" = "VOLUME"))


# create counts per study
# Aim is to add number of participants  in image

counts_df <- meta_unique %>%
  group_by(study) %>%
  summarise(
    n = n(),
    male = sum(sex == "male"),
    female = sum(sex == "female"),
    .groups = "drop"
  )


# for the summarised image

sum_df <- meta_unique %>%
  summarise(
    n = n(),
    male = sum(sex == "male"),
    female = sum(sex == "female")
  )


# To help set in-image text
max_count <- ggplot_build(
  ggplot(meta_unique, aes(x = age)) +
    geom_histogram(binwidth = 5)
)$data[[1]]$count %>% max()


# Plot the distribution of all participants in one image
all <- ggplot(meta_unique, aes(x = age, fill = sex)) +
  geom_histogram(position = position_dodge(width = 5),
                 alpha = 0.5,
                 binwidth = 5,
                 color = "black") +
  scale_fill_manual(
    values = c(
      "male" = colors[7],
      "female" = colors[3]
    )
  ) +
  geom_text(
    data = sum_df, aes(x = min(meta_unique$age), y= 0.5 * max_count,
                       label = paste0("n = ", n,
                                      "\nMales = ", male,
                                      "\nFemales = ", female)),
    inherit.aes = F,
    hjust = 0.001,
    vjust = 0.01,
    size = 2,
    fontface= "bold.italic"
  )+
  # facet_wrap(~ study, ncol = 2) +
  theme_minimal(base_size = 10) +
  labs(
    title = "Age distribution of all participants",
    x = "Age",
    y = NULL
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(face= "bold"),
    legend.title = element_blank(),
    legend.text = element_text(size = 10, face = "bold"), 
    axis.text.y = element_text(size = 10, face= "bold"),
    axis.text.x = element_text(size = 10, face= "bold"),
    axis.title = element_text(size = 10, face= "bold")
  )



## Make the per study plot

by_study <- ggplot(meta_unique, aes(x = age, fill = sex)) +
  geom_histogram(position = "dodge",
                 alpha = 0.5,
                 binwidth = 5,
                 color = "black") +
  scale_fill_manual(
    values = c(
      "male" = colors[7],
      "female" = colors[3]
    )
  ) +
  facet_wrap(~ study, ncol = 3, scales = "free") +
  geom_text(
    data = counts_df, aes(x = min(meta_unique$age), y= 0.5 * max_count,
                          label = paste0("n = ", n,
                                         "\nMales = ", male,
                                         "\nFemales = ", female)),
    inherit.aes = F,
    hjust = 0.1,
    vjust = 1.5,
    size = 2,
    fontface= "bold.italic"
  )+
  theme_minimal(base_size = 10) +
  labs(
    title = " Distribution of Participants across all 5 studies",
    x = "Age",
    y = "Number of participants"
  ) +
  
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_text(face= "bold"),
     legend.title = element_blank(),
     legend.text = element_text(size = 10, face = "bold"), 
   # legend.position = "none",
    axis.text.y = element_text(size = 10, face= "bold"),
    axis.text.x = element_text(size = 10, face= "bold"),
    axis.title = element_text(size = 10, face= "bold"),
   legend.position = "none"
  )




# Plot the distribution of splicing efficiency across all introns



# Distribution of splicing efficiency values across the dataset
long_splice <- all_splice_df %>% 
  pivot_longer(names_to = "seq_sample_id",
               values_to = "SE",
               cols = -(transcript_ID))

dist_df <- long_splice %>%
  filter(!is.na(SE)) %>%
  mutate(
    SE_range = case_when(
      SE <= 0.25 ~ "SE from 0.00-0.25",
      SE <= 0.49 ~ "SE from 0.26-0.49",
      SE <= 0.75 ~ "SE from 0.50-0.75",
      TRUE       ~ "SE from 0.76-1.00"
    ),
    SE_range = factor(SE_range,
                      levels = c("SE from 0.00-0.25", "SE from 0.26-0.49",
                                 "SE from 0.50-0.75", "SE from 0.76-1.00"))
  ) %>%
  count(SE, SE_range) %>%
  mutate(percent = n / sum(n) * 100,
         label = ifelse(SE %in% c(0, 1),
                        paste0(round(percent, 1), "%"),
                        NA)) 


introns_distribution <-   ggplot(dist_df,
                                 aes(x = factor(SE, levels = sort(unique(SE))),
                                     y = percent,
                                     fill = case_when(
                                       SE == 1 ~ "one",
                                       SE == 0 ~ "zero",
                                       TRUE    ~ "other"
                                     ))) +
  geom_col() +
  
  # geom_text(
  #   data = subset(dist_df, SE %in% c(0, 1)),
  #   aes(label = paste0(round(percent, 1), "%")),
  #   vjust = 0.5,
  #   hjust=0.5,
  #   size = 5,
  #   fontface = "bold"
  # ) +
  # Label for SE = 0
  geom_text(
    data = subset(dist_df, SE == 0),
    aes(label = paste0(round(percent, 1), "%")),
    vjust = 1.5,
    hjust = -0.2   # adjust differently
  ) +
  
  # Label for SE = 1
  geom_text(
    data = subset(dist_df, SE == 1),
    aes(label = paste0(round(percent, 1), "%")),
    vjust = 1.5,
    hjust = 1.2    # adjust differently
  )+

  facet_wrap(~SE_range, scales = "free") +
  
  scale_fill_manual(
    values = c(
      "zero"  = colors[1],  
      "one"   = colors[6],   
      "other" = colors[11]
    ),
    guide = "none"
  ) +
  
  scale_x_discrete(breaks = seq(0, 1, by = 0.1)) +
  theme_minimal(base_size = 10) +
  labs(
    x = "Splicing Efficiency (SE)",
    y = "Percentage",
    title = "Distribution of Splicing Efficiency Values across the pooled dataset"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_blank(), # removes facet titles
    legend.title = element_blank(),
    legend.text = element_text(size = 10, face = "bold"), 
    axis.text.y = element_text(size = 10, face= "bold"),
   # axis.text.y = element_blank(),  remove y axis title
    axis.text.x = element_text(angle = 90, size = 10, face= "bold"),
    axis.title = element_text(size = 10, face= "bold"),
   axis.title.y = element_blank(),
    panel.spacing.y = unit(2, "lines"),
    panel.grid = element_blank(), # removes grid lines 
    
    panel.border = element_rect( # add a subtle border
      colour = "grey80",   
      fill = NA,
      linewidth = 0.5
    ))



## The RELIEF data



Relief_full_meta <- readRDS("data/Relief_metadata.RDS")%>%
  dplyr::select(study, participant, sex, time, seq_sample_id, age) %>%
  mutate(sex = factor(sex, levels = c("female", "male")),
         time = factor(time, levels = c("PreExc", "PostExc")),
         age_group = case_when(age < 40 ~ "Young",
                               age > 40 ~ "Old"),
         age_group = factor(age_group, levels = c("Young", "Old")))


Relief_full_splice <- readRDS("data/Relief_splicing_data.RDS")  %>%
  drop_na()



# REORDER THE SEQUENCE ID TO MATCH BOTH DATAFRAMMES
splice_reordered <- Relief_full_splice[, c("transcript_ID",Relief_full_meta$seq_sample_id)] 




long_relief <- splice_reordered %>% 
  pivot_longer(names_to = "seq_sample_id",
               values_to = "SE",
               cols = -(transcript_ID))

dist_df_relief <- long_relief %>%
  filter(!is.na(SE)) %>%
  mutate(
    SE_range = case_when(
      SE <= 0.25 ~ "SE from 0.00-0.25",
      SE <= 0.49 ~ "SE from 0.26-0.49",
      SE <= 0.75 ~ "SE from 0.50-0.75",
      TRUE       ~ "SE from 0.76-1.00"
    ),
    SE_range = factor(SE_range,
                      levels = c("SE from 0.00-0.25", "SE from 0.26-0.49",
                                 "SE from 0.50-0.75", "SE from 0.76-1.00"))
  ) %>%
  count(SE, SE_range) %>%
  mutate(percent = n / sum(n) * 100,
         label = ifelse(SE %in% c(0, 1),
                        paste0(round(percent, 1), "%"),
                        NA)) 


relief_distribution <-   ggplot(dist_df_relief,
                                 aes(x = factor(SE, levels = sort(unique(SE))),
                                     y = percent,
                                     fill = case_when(
                                       SE == 1 ~ "one",
                                       SE == 0 ~ "zero",
                                       TRUE    ~ "other"
                                     ))) +
  geom_col() +
  
  # geom_text(
  #   data = subset(dist_df_relief, SE %in% c(0, 1)),
  #   aes(label = paste0(round(percent, 1), "%")),
  #   vjust = 1.5,
  #   hjust=0.5,
  #   size = 4,
  #   fontface = "bold"
  # ) +
  # Label for SE = 0
  geom_text(
    data = subset(dist_df_relief, SE == 0),
    aes(label = paste0(round(percent, 1), "%")),
    vjust = 1.5,
    hjust = -0.2   # adjust differently
  ) +
  
  # Label for SE = 1
  geom_text(
    data = subset(dist_df_relief, SE == 1),
    aes(label = paste0(round(percent, 1), "%")),
    vjust = 1.5,
    hjust = 1.2    # adjust differently
  ) +

  
  facet_wrap(~SE_range, scales = "free") +
  
  scale_fill_manual(
    values = c(
      "zero"  = colors[1],  
      "one"   = colors[6],   
      "other" = colors[11]
    ),
    guide = "none"
  ) +
  
  scale_x_discrete(breaks = seq(0, 1, by = 0.1)) +
  theme_minimal(base_size = 10) +
  labs(
    x = "Splicing Efficiency (SE)",
    y = "Percentage",
    title = "Distribution of Splicing Efficiency values across the RELIEF cohort"
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.text = element_blank(), # removes facet titles
    legend.title = element_blank(),
    legend.text = element_text(size = 10, face = "bold"), 
    axis.text.y = element_text(size = 10, face= "bold"),
    axis.text.x = element_text(angle = 90, size = 10, face= "bold"),
    axis.title = element_text(size = 10, face= "bold"),
    panel.spacing.y = unit(2, "lines"),
    panel.grid = element_blank(), # removes grid lines 

 panel.border = element_rect( # add a subtle border
   colour = "grey80",
   fill = NA,
   linewidth = 0.5
   )
) 



bubble_img <- png::readPNG("Figures/Bubble_chart.PNG")
image_grob <- grid::rasterGrob(bubble_img, interpolate = TRUE)

image_plot <- wrap_elements(image_grob)




# explore the introns completely retained, or perfectly spliced across all samples

long_splice <- all_splice_df %>% 
  pivot_longer(names_to = "seq_sample_id",
               values_to = "SE",
               cols = -(transcript_ID))

dist_df <- long_splice %>%
  filter(!is.na(SE)) %>%
  mutate(
    SE_range = case_when(
      SE <= 0.25 ~ "SE from 0.00-0.25",
      SE <= 0.49 ~ "SE from 0.26-0.49",
      SE <= 0.75 ~ "SE from 0.50-0.75",
      TRUE       ~ "SE from 0.76-1.00"
    ),
    SE_range = factor(SE_range,
                      levels = c("SE from 0.00-0.25", "SE from 0.26-0.49",
                                 "SE from 0.50-0.75", "SE from 0.76-1.00"))
  ) %>%
  count(SE, SE_range) %>%
  mutate(percent = n / sum(n) * 100,
         label = ifelse(SE %in% c(0, 1),
                        paste0(round(percent, 1), "%"),
                        NA)) 

all_zeros <- long_splice %>%
  filter(SE == 0) %>%
  pivot_wider(names_from = seq_sample_id,
              values_from = SE)

# extract transcript IDs that were zeros all through
zero_introns <- all_splice_df %>%
  filter(if_all(-transcript_ID, ~ . == 0)) %>%
  dplyr::select(transcript_ID) %>%
  separate(transcript_ID, into = c("transcript_ID", "intron_ID", "chr"),
           sep = "_") %>%
  inner_join(gene_annotation,
             by = c("transcript_ID" = "ensembl_transcript_id_version"))  %>%
  mutate(
    gene_label  = ifelse(
      is.na(external_gene_name) | external_gene_name == "",
      ensembl_gene_id, external_gene_name
    ),
    gene_intron = paste(gene_label, intron_ID, sep = " : ")
  )

zero_introns_relief <- splice_reordered%>%
  filter(if_all(-transcript_ID, ~ . == 0)) %>%
  dplyr::select(transcript_ID) %>%
  separate(transcript_ID, into = c("transcript_ID", "intron_ID", "chr"),
           sep = "_") %>%
  inner_join(gene_annotation,
             by = c("transcript_ID" = "ensembl_transcript_id_version")) %>%
  mutate(
    gene_label  = ifelse(
      is.na(external_gene_name) | external_gene_name == "",
      ensembl_gene_id, external_gene_name
    ),
    gene_intron = paste(gene_label, intron_ID, sep = " : ")
  )


one_introns <- all_splice_df %>%
  filter(if_all(-transcript_ID, ~ . == 1)) %>%
  dplyr::select(transcript_ID) %>%
  separate(transcript_ID, into = c("transcript_ID", "intron_ID", "chr"),
           sep = "_") %>%
  inner_join(gene_annotation,
             by = c("transcript_ID" = "ensembl_transcript_id_version")) %>%
  mutate(
    gene_label  = ifelse(
      is.na(external_gene_name) | external_gene_name == "",
      ensembl_gene_id, external_gene_name
    ),
    gene_intron = paste(gene_label, intron_ID, sep = " : ")
  )




ego_ones <- enrichGO(gene =  unique(one_introns$external_gene_name),
                     keyType = "SYMBOL",
                     universe = gene_exp_df$gene_name,
                     OrgDb = org.Hs.eg.db, 
                     ont = "BP", 
                     pAdjustMethod = "BH", 
                     qvalueCutoff = 0.05, 
                     readable = T)


## Output results from GO analysis to a table
cluster_ones <- data.frame(ego_ones)





go_ones <- dotplot(ego_ones,
                   showCategory = 6,
                   font.size = 8, title = "Enriched biological processes in genes containing introns perfectly spliced acrosss all samples") +
  theme(axis.text = element_text(size = 10), axis.text.y = element_text(size = 10), axis.title.x = element_text(size = 10),
        plot.title = element_text(hjust = 0.5, size = 10, face = "bold"),
        legend.text = element_text(size = 10, face = "bold"),
        legend.title = element_text(size = 10, face = "bold"))

# ggsave("SVG_files/gene_ontology_ones.svg", plot = go_ones, width = 16, height = 15)

# Repeat for RELIEF

one_introns_relief <- splice_reordered %>%
  filter(if_all(-transcript_ID, ~ . == 1)) %>%
  dplyr::select(transcript_ID) %>%
  separate(transcript_ID, into = c("transcript_ID", "intron_ID", "chr"),
           sep = "_") %>%
  inner_join(gene_annotation,
             by = c("transcript_ID" = "ensembl_transcript_id_version")) %>%
  mutate(
    gene_label  = ifelse(
      is.na(external_gene_name) | external_gene_name == "",
      ensembl_gene_id, external_gene_name
    ),
    gene_intron = paste(gene_label, intron_ID, sep = " : ")
  )



ego_ones_relief <- enrichGO(gene =  unique(one_introns_relief$external_gene_name),
                            keyType = "SYMBOL",
                            universe = gene_exp_df$gene_name,
                            OrgDb = org.Hs.eg.db, 
                            ont = "BP", 
                            pAdjustMethod = "BH", 
                            qvalueCutoff = 0.05, 
                            readable = T)


## Output results from GO analysis to a table
cluster_ones_relief <- data.frame(ego_ones_relief)




go_ones_relief <- dotplot(ego_ones_relief,
                          showCategory = 6,
                          font.size = 10, title = "Enriched biological processes in genes containing introns perfectly spliced in Relief samples") +
  theme(axis.text = element_text(size = 10), axis.text.y = element_text(size = 10), axis.title.x = element_text(size = 10),
        plot.title = element_text(hjust = 0.5, size = 10, face = "bold"),
        legend.text = element_text(size = 10, face = "bold"),
        legend.title = element_text(size = 10, face = "bold") )

#ggsave("SVG_files/gene_ontology_ones_RELIEF.svg", plot = go_ones_relief, width = 16, height = 15)


# 
 Figure_1 <- ( by_study + all + image_plot) /(relief_distribution + introns_distribution) /
   (go_ones_relief + go_ones)
# 
# 
Figure_1 +
  plot_annotation(tag_levels = "A") &
  #plot_layout(widths = c(1.1, 1))
  theme(
    plot.tag = element_text(size = 10, face = "bold")
    ,
    plot.tag.position = c(0.08, 0.98)
  )

```

## Figure 2 

Visualising the association between aging and intron splicing efficiency in the primary analysis, and the trajectories of topmost differentially spliced introns in the secondary analysis

```{r}
#| echo: false

# The marginal effects predictions were built into the summary function and thus
# Extracting the data will follow the traditional seqwrap method
Relief_results <- seqwrap_summarise(Relief_model)

Relief_results <- Relief_results$summaries



# Only extracts the contrasts of interest
contrasts_of_interest <- c("train_young", "train_old", 
                           "age_effect_pre", "age_effect_post", 
                           "train_old_young")

relief_contrasts <- Relief_results %>%
  filter(hypothesis %in% contrasts_of_interest) %>%
  group_by( hypothesis, component) %>%
  mutate(adj.p = p.adjust(p.value, method = "fdr"),
         # recode the components and hypothesis for easy understanding
         component = recode(component, 
                                   "response"    = "Overall retention",
                                   "zprob"       = "Probability of Perfect splicing",
                                   "conditional" = "Degree of retention"),
         hypothesis = recode(hypothesis, "train_young"     = "Exercise effect among young participants",
                                        "train_old"       = "Exercise effect among older participants",
                                        "age_effect_pre"  = "Age effect at baseline",
                                        "age_effect_post" = "Age effect postexercise",
                                        "train_old_young" = "Interaction effect of training and aging")) %>%
  ungroup() %>%
  mutate(
    sig           = adj.p <= 0.05,
    sig_ci        = conf.low > 0 | conf.high < 0,
    neg_log10_fdr = -log10(adj.p),
    rank_score    = -log10(adj.p) * abs(estimate),
    effect        = case_when(
      conf.high < 0 & sig ~ "Improved SE",
      conf.low  > 0 & sig ~ "Reduced SE",
      TRUE             ~ "No effect"
    )
  ) %>%
  annotate_introns(gene_annotation, intron_length, flip = TRUE) 


# Which introns were differentially spliced among the aged at baseline
aging_baseline <- relief_contrasts %>%
  filter(hypothesis ==  "Age effect at baseline") %>%
  filter(component != "Overall retention")



top_9_labels <- aging_baseline %>%
  filter(sig) %>%
  group_by(component) %>%
  mutate(rank_score = abs(estimate) * -log10(adj.p)) %>% 
  arrange(desc(rank_score), adj.p, desc(abs(estimate))) %>%
  slice_max(rank_score, n = 9, with_ties = FALSE) %>%
  ungroup()

# Count summary per facet
facet_sum <- aging_baseline %>%
  group_by( component) %>%
  summarise(
    n_total  = n(),
    n_sig    = sum(sig),
    perc_sig = round(100 * n_sig / n_total, 1),
    x        = Inf,
    y        = Inf,
    label    = paste0("ds: ", n_sig, "/", n_total,
                      " (", perc_sig, "%)"),
    .groups  = "drop"
  )

volcano_aging_baseline <- ggplot(aging_baseline,
       aes(estimate, neg_log10_fdr, colour = effect)) +
  geom_point(alpha = 0.6, size = 2.5) +
  geom_text_repel(
    data        = top_9_labels,
    aes(label   = gene_intron),
    size        = 4,
    max.overlaps = Inf
  ) +
  geom_text(
    data         = facet_sum,
    aes(x = x, y = y, label = label),
    inherit.aes  = FALSE,
    hjust        = 2.9,
    vjust        = 2.9,
    size         = 4,
    fontface     = "bold"
  ) +
  geom_hline(yintercept = -log10(0.05),
             linetype = "dashed", colour = "grey50") +
  geom_vline(xintercept = 0,
             linetype = "dashed", colour = "grey50") +
  scale_colour_manual(values = effect_colors) +
  scale_y_continuous(limits = c(0, 16)) +  # adjust 20 to a sensible cap
  coord_cartesian(clip = "off")+
  facet_wrap(~ component, scales = "free") +
  labs(
    title    = "Introns affected by aging at baseline",
      subtitle = "Top 9 differentially spliced introns annotated",
    x        = "Effect size (SE scale)",
    y        = expression(-log[10](FDR)),
    colour   = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 10),
    strip.text = element_text(face= "bold", size = 12),
    legend.title = element_blank(),
    legend.text = element_text(size = 12, face = "bold"), 
    # legend.position = "none",
    axis.text.y = element_text(size = 12, face= "bold"),
    axis.text.x = element_text(size = 12, face= "bold"),
    axis.title = element_text(size = 12, face= "bold")
  )




# extract the trajecories of the top 12


ds_intron_aging <- sig_baseline %>%
  #arrange(desc(abs(estimate))) %>%
  mutate(rank_score = abs(estimate) * -log10(adj.p)) %>% 
  arrange(desc(rank_score), adj.p, desc(abs(estimate))) %>%
  slice_head(n = 59) # this is the number at which 12 common introns were in both primary and secondary



aged_intersect <- zi_predictions %>%
  filter(target %in% ds_intron_aging$target) %>% # extract top 9
  mutate(
    SE       = 1 - estimate,
    CI_low   = 1 - conf.high,
    CI_high  = 1 - conf.low,
    real_age = scaled_age * (age_max - age_min) + age_min
  ) %>%
  left_join(
    (zi_age_slopes_fdr  %>%
       dplyr::select(target, gene_intron) %>% distinct()),
    by = "target"
  ) %>%
  filter(time == "PreExc")

length(unique(aged_intersect$gene_intron))

traj_aged <- aged_intersect %>%
  ggplot(aes(x = real_age, y = SE, colour =  gene_intron)) +
  geom_ribbon(aes(ymin = CI_low, ymax = CI_high),
              alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.9) +
  facet_wrap(~ gene_intron, scales = "free_y", ncol = 3) +
  labs(
    title    = "Age-related trajectory of the top 12 ds introns reflected in secondary analysis",
    x        = "Age (years)",
    y        = "Splicing efficiency",
    colour   = NULL, fill = NULL
  ) +
  theme_minimal(base_size = 16) +
  theme(
    plot.title      = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle   = element_text(hjust = 0.5),
    strip.text      = element_text(face = "bold", size = 14)
    
    
  )


plot_global_trajectory <- zi_predictions %>%
  mutate(
    SE       = 1 - estimate,
    CI_low   = 1 - conf.high,
    CI_high  = 1 - conf.low,
    real_age = scaled_age * (age_max - age_min) + age_min
  ) %>%
  group_by(real_age, time) %>%
  summarise(
    mean_SE = mean(SE),
    CI_low  = mean(CI_low),
    CI_high = mean(CI_high),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = real_age, y = mean_SE, colour = time, fill = time)) +
  geom_ribbon(aes(ymin = CI_low, ymax = CI_high),
              alpha = 0.15, colour = NA) +
  geom_line(linewidth = 0.9) +
  geom_smooth(method = "lm", se = FALSE,
              linetype = "dashed", linewidth = 0.6) +
  scale_colour_manual(values = c("PreExc" = colors[5],
                                 "PostExc" = colors[1])) +
  scale_fill_manual(values   = c("PreExc" = colors[5],
                                 "PostExc" = colors[1])) +
  labs(
    title    = "Global splicing efficiency trajectory across age",
    subtitle = "Solid = spline fit, dashed = linear reference",
    x        = "Age (years)",
    y        = "Mean splicing efficiency",
    colour   = NULL, fill = NULL
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
     plot.subtitle = element_text(hjust = 0.5, size = 13),
    strip.text = element_text(face= "bold"),
    legend.title = element_blank(),
    legend.text = element_text(size = 12), 
    # legend.position = ,
    axis.text.y = element_text(size = 12, face= "bold"),
    axis.text.x = element_text(size = 12, face= "bold"),
    axis.title = element_text(size = 12, face= "bold"),
    panel.grid = element_blank(), 
  )
volcano_aging_baseline + traj_aged +  plot_annotation(tag_levels = 'A')


# ggsave("Figures/Figure_2.png",  width = 26, height = 15, dpi = 600)
```

The distribution of splicing efficiency values across all introns

```{r}
#| echo: false


```

### The RELIEF data

```{r}
#| echo: false



```
