---
title: "Beta regression simulations"
format: typst
bibliography: resources/extra-bibliography.bib
bibliographystyle: "the-lancet"
editor_options: 
  chunk_output_type: console
---

## The Beta distribution

The Beta distribution is parameterized by $\alpha$ and $\beta$, called shape parameters. These are positive numbers and related to the central tendency $\mu$ and inverse dispersion (or precision [@douma2019]), $\phi$ of the distribution:

$$
\begin{align}
\alpha &= \mu \times \phi \\
\beta &= (1 - \mu) \times \phi
\end{align}
$$

$\mu$ is modelled on the logit scale where $\eta$ is the linear predictor. To get $\mu$ from $\eta$ we apply the inverse logit transformation:

$$\mu = \frac{\operatorname{exp}(\eta)}{1+\operatorname{exp}(\eta)}$$

The inverse dispersion parameter $\phi$ is modelled on the log scale:

$$\phi = \operatorname{exp}(1/\sigma)$$ To simulate values from the beta distribution we can use `rbeta()`, e.g.:

```{r}
#| message: false
#| warning: false


library(tidyverse)

y <- rbeta(100, 2, 5)


data.frame(y) |>
  ggplot(aes(y)) + geom_histogram()

```

The mean of the Beta distribution ($\mu$) is related to the linear predictor, $\eta$, as shown below.

```{r}
eta <- seq(from = -6, to = 6, by = 0.1)

data.frame(eta, mu = plogis(eta)) |>
  ggplot(aes(eta, mu)) + geom_line()
```

## Simulating beta-distributed data

In our study we want to explore the effect of age (a continuous variable) and training status (a factor variable) on the proportion of retained introns (IR; $\text{IR} \in [0,1]$).

$$
\begin{align}
\text{IR} &\sim \text{Beta}(\alpha, \beta) \\
\alpha &= \mu \times \phi \\
\beta &= (1 - \mu) \times \phi \\
\mu &= \frac{\operatorname{exp}(\eta)}{1+\operatorname{exp}(\eta)} \\
\phi &= \operatorname{exp}(1/\sigma) \\
\eta &= \beta_0 + \beta_1 \times \text{age} + \beta_2 \times \text{training} \\
\sigma &= 2
\end{align}
$$

A simulation function is constructed as

```{r}
ir_sim1 <- function(beta0, beta1, beta2, sigma, age, training) {
  
  n <- length(age)
  eta <- beta0 + beta1 * age + beta2 * training
  
  phi <- exp(sigma)
  mu <- exp(eta) / (1 + exp(eta))
  
  alpha <- mu * phi
  beta <- (1 - mu) * phi
  
  y <- rbeta(n, alpha, beta)
  
  y
}
```

The raw data from the simulation function produces outcome on the scale $y\in[0,1]$ based on the parameters for the linear predictor of $\eta$.

```{r}
## Test the function 



Age <- runif(500)
Training <- rep(c(0,1), each = 250)


y <- ir_sim1(beta0 = 3, beta1 = -6, beta2 = 4, sigma = 0.001,  age = Age, training = Training)

dat0 <- data.frame(Age, Training, y) 

dat0 |>
  ggplot(aes(Age, y, color = as.factor(Training))) + 
  geom_point()


```

The `rbeta` function returns values in the interval $[0,1]$ including boudaries (0, and 1). However, modelling with the `beta_family` distribution does not allow for boundary values (0,1). We have to compress the data prior to fitting.

```{r}


compress <- function(x, e = 10^-4, additative = TRUE) {
  
  n <- length(x)
  
  if(additative) {
    xout <- x * (1 - 2*e) + e
  } else {
    xout <- (x * (n - 1) + 0.5) / n
  }
  
  xout
  
}


dat0$y2 <- compress(dat0$y)



library(glmmTMB)


m <- glmmTMB(y2 ~ Age + Training, data = dat0, 
             family = "beta_family")


summary(m)



```

The simulation function together with compression seems to return a data sets that we are able to model. In the construction of the parameters of the Beta distribution ($\alpha$ and $\beta$), the dispersion parameter $\phi$ is of importance. We will likely not get the coefficients from $\eta$ back from the model due to $\phi$. To explore the effect of $\phi$:

```{r}

sigmas <- seq(from = 0.1, to = 10, by = 0.7)
Age <- runif(500)
Training <- rep(c(0,1), each = 250)

results <- list()


for(i in seq_along(sigmas)) {
  
  # Simulate data
  y <- ir_sim1(beta0 = 3, beta1 = -6, beta2 = 4, sigma = sigmas[i],  age = Age, training = Training)

  dat_sim <- data.frame(Age, Training, y) 
  
  dat_sim$y2 <- compress(dat_sim$y)
  
  # Fit model
  m <- glmmTMB(y2 ~ Age + Training, data = dat_sim, 
             family = "beta_family")

  results[[i]] <- data.frame(param = c("b0", "b1", "b2"), 
             coef = fixef(m)$cond,
             sigma = sigmas[i],
             row.names = NULL)
  
}

results <- bind_rows(results) 


results |>
  mutate(eta_beta = rep(c('beta[0]*" = 3"',
                          'beta[1]*" = -6"', 
                          'beta[2]*" = 4"'), length.out = length(sigmas) * 3)) |>
  
  ggplot(aes(sigma, coef)) + geom_point() + 
  facet_wrap(~ eta_beta, labeller = label_parsed)
  



```

As $\phi$ increases, the precision increases. This means that we should think about $\sigma$ in the simulation as being specified as $1/\sigma$. A large $\sigma$ will result in small precision (and larger variation in the coefficients).

## Multilevel Beta

How can we interpret coefficients in the multilevel version of the Beta model?

Interpretation of our data will benefit from interpretation using a multilevel model. Data comes from a number of different studies, with multiple samples from participants in each study.

```{r}

# beta0, global intercept
# beta1, global slope for age
# beta2, global training effect
# sigma0/sigma1/sigma2, SD of global effects between studies
# sigma_participant, SD of between participant intercepts
# sigma, "residual" variation should be entered as precision 1/sigma
# age/training/study/participants, design variables
#
#
# returns a vector of simulated IR values based on the design variables.
ir_sim2 <- function(beta0 = 3, 
                    beta1 = -3, 
                    beta2 = 3, 
                    sigma0 = 1, 
                    sigma1 = 1, 
                    sigma2 = 1, 
                    sigma = 1/0.5, 
                    sigma_participant = 1,
                    age = c(1, 2, 1, 2), 
                    training = c(0,0,0,0), 
                    study = c(1, 2, 3, 4), 
                    participant = c(1, 2, 4, 5)) {
  
  # Save n observations
  n <- length(participant)
  # Check that all design variables are of equal length
  if(length(unique(lengths(list(participant, study, training, age)))) != 1) stop("All design variables (age, training, study, participant) must be of the same length.") 
   
  
  # List all studies
  studies <- unique(study)
  
  ## Draw intercepts per study
  beta0s <- rnorm(length(studies), beta0, sigma0)
  
  ## Draw age-slopes per study
  beta1s <- rnorm(length(studies), beta1, sigma1)
  
  ## Draw training slopes per study
  beta2s <- rnorm(length(studies), beta2, sigma2)
  
  ## List all participants
  participants <- unique(paste0(participant, "_", study))
  
  ## Draw intercept per participant
  b0s <- rnorm(length(participants), 0, sigma_participant) 
  
  # Combine all drawn values
  
  partdat <- data.frame(participant_study = participants, 
                     b0 = b0s) |>
    tidyr::separate(participant_study, 
                    into = c("participant", "study"), 
                    sep = "_")
  
  studydat <- data.frame(study = as.character(studies), 
                      beta0 = beta0s, 
                      beta1 = beta1s, 
                      beta2 = beta2s)
  
  eta <- partdat |>
    dplyr::inner_join(studydat, by = "study") |>
    dplyr::inner_join(data.frame(age, 
                                 training, 
                                 study = as.character(study), 
                                 participant = as.character(participant))) |>
    dplyr::mutate(eta = beta0 + beta1 * age + beta2 * training + b0) |>
    dplyr::pull(eta)
  
  phi <- exp(sigma)
  mu <- exp(eta) / (1 + exp(eta))
  
  alpha <- mu * phi
  beta <- (1 - mu) * phi
  
  y <- rbeta(n, alpha, beta)
  
  return(y)
  
}


```

### Multilevel Beta regression with simulated data

```{r}

design <- rbind(
  # Study 1
  data.frame(study = 1, 
             participant = rep(1:20, each = 2), 
             age = rep(runif(20, 0.5, 1), each = 2), 
             training = c(0,1)), 
  # Study 2
   data.frame(study = 2, 
             participant = rep(1:20, each = 2), 
             age = rep(runif(20, 0.1, 0.6), each = 2), 
             training = c(0,1)), 
  # Study 3
  data.frame(study = 3, 
             participant = rep(1:20, each = 2), 
             age = rep(runif(20, 0.1, 0.9), each = 2), 
             training = c(0,1)), 
  # Study 4
    data.frame(study = 4, 
             participant = rep(1:20, each = 2), 
             age = rep(runif(20, 0.01, 0.3), each = 2), 
             training = c(0,1)), 
  # Study 5
     data.frame(study = 5, 
             participant = rep(1:20, each = 2), 
             age = rep(runif(20, 0.01, 0.99), each = 2), 
             training = c(0,1))
  
)


```

Next we simulate (and compress) the data.

```{r}

y <- ir_sim2(beta0 = 3, beta1 = -1.5, beta2 = 2, 
             sigma0 = 2, sigma1 = 1, sigma2 = 2,
        sigma = 1/0.5, 
        sigma_participant = 0.66, 
        age = design$age, 
        training = design$training, 
        participant = design$participant, 
        study = design$study)


dat1 <- design |>
  mutate(participant = paste0(study, "_", participant))

dat1$y <- compress(y)


```

```{r}

dat1 |>
  ggplot(aes(age, y, 
             color = as.factor(study))) + 
  geom_smooth(se = FALSE) +
  geom_point()


```

```{r}

m <- glmmTMB(y ~ age * training + 
               (1|study) + (1|participant), 
             data = dat1, 
             family = beta_family() )


summary(m)
VarCorr(m)

```
