---
title: "Model_exploration"
format: html
editor: visual
execute: 
  warning: false
  message: false
  code-summary: "Code"
  code-fold: true
  echo: true
  fig-align: center
  output: true
editor_options: 
  chunk_output_type: inline
---

## Model exploration

This explores the model and visualises its output

Load the relevant libraries and files

```{r}
library(dplyr)
library(tidyverse)
library(gridExtra)
library(ggpubr)
library(cowplot)
library(ggridges)
library(UpSetR)
library(ggplot2)
library(ggplotify)
library(clusterProfiler)
library(org.Hs.eg.db)
library(magick)


# Load the model summary
model_summary <- readRDS("data/RT_model_summary.RDS")

# Load original splicing data 
all_splicing <- readRDS("data/all_splice.RDS") %>%
  drop_na()
all_metadata <- readRDS("data/all_full_metadata.RDS")


# Load the batch-corrected gene expression data
gene_exp_df <- readRDS("data_new/gene_counts/batch_corrected_genecounts.RDS") #%>%
  # remove the version number to match gene_id
 # mutate(gene_id = gsub("\\..*", "",  gene_id))



# Load the gene annotation file
gene_annotation <- readRDS("data/ensembl_gene_annotation.RDS")

# Load one file from which we will extract intron length
# This is valid as only introns quantified in all samples were included in the analyses
intron_length <- readr::read_tsv("data_new/Alpha_Omega_SpliceQ_outputs/A_102.tsv") %>%

distinct(across(6:ncol(.)), .keep_all = T) %>% # Removes duplicates based on columns 6 to end
  mutate(transcript_ID = paste0(transcript_ID, "_", intron_ID, "_", chr),
    intron_length = abs((sj3start - sj5end) + 1) ) %>% # Ensures positive length regardless of strand
  dplyr::select(transcript_ID, intron_length)


```

Extract the relevant columns

```{r}

# extract the model summaries.. 
RT_model_summary<- model_summary$summaries %>%
  # The summaries contain both the model summaries and the predictions
  # select only the summary outputs
  dplyr::select(coef, target,Estimate, Pr...z..) %>%
  # select the model results of those affected by RT alone or interaction with aging
  filter(coef == "timePostExc" | coef == "scaled_age:timePostExc" | coef == "scaled_age") %>%
  mutate(adj.p = p.adjust( Pr...z.., method = "fdr"),
         log2fc = Estimate/log(2),
         fcthreshold = if_else(abs(log2fc) > 0.5, "s", "ns"),
         odds_ratio = exp(Estimate)) %>%
  inner_join(intron_length, by = c("target" = "transcript_ID"))




# Extract the prdictions

# From the model summary, extract those that are predictions
RT_predictions <- model_summary$summaries%>%
  dplyr::select(scaled_age, time, fit, target) %>%
  drop_na() %>%
  # Extract the transcript_id from the target
  mutate(transcript_ID = str_split(target, "_",simplify= T) [,1])



# Merge the summary and predictions into one clearner dataframe
# merge dataset
RT_merged <- RT_predictions %>%
  inner_join(RT_model_summary, by = "target") %>%
  drop_na() %>%
  # based on model estimate and p values, group the introns into how they are affected by aging
  mutate(effect = case_when(Estimate > 0 & Pr...z.. <= 0.05 ~ "Improved SE", 
                            Estimate < 0 & Pr...z.. <= 0.05 ~ "Reduced SE" ,
                            Estimate < 0 & Pr...z.. > 0.05 ~ "No effect",
                            Estimate > 0 & Pr...z.. > 0.05 ~ "No effect")) %>%
  inner_join(gene_annotation, by= c("transcript_ID" = "ensembl_transcript_id_version"))

# length(unique(RT_merged$target))
saveRDS(RT_merged, "data/RT_model_df.RDS")

```

## Basic visualisation

```{r}
# plot the distribution of gene biotypes of the genec containing the introns
intron_distribution <- RT_merged %>%
  distinct(target, transcript_biotype, .keep_all = T) %>%
  ggplot(aes(transcript_biotype, fill = transcript_biotype))+
  geom_bar()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        plot.title = element_text(hjust = 0.5),
        legend.position = "none")+ # no legend
  stat_count(geom = "Text", aes(label = ..count..), vjust = -0.08) +
  ggtitle("Distribution of biotypes of genes containing the introns") +
  ylab("Number of introns per biotype") 
  
intron_distribution
```

```{r}

# Plot the gene biotype to check if the distribution of biotypes also fit that of those containing introns
gene_distribution <- gene_annotation %>%
  distinct(external_gene_name, .keep_all = T) %>%
  inner_join((gene_exp_df %>%
                dplyr::select(gene_name) ), by = c("external_gene_name" = "gene_name")) %>%
#  distinct(gene_id, transcript_biotype, .keep_all = T)%>%
  ggplot(aes(transcript_biotype, fill = transcript_biotype))+
  geom_bar()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1),
        plot.title = element_text(hjust = 0.5),
        legend.position = "none")+
  stat_count(geom = "Text", aes(label = ..count..), vjust = -0.1) +
  ggtitle("Distribution of gene biotypes in the gene expression data") +
  ylab("Number of genes per biotype")
gene_distribution
```

## Aging-associated effects

```{r}

# Explore the effect of aging alone and extract the number of introns per category
Aging_effect <- RT_merged %>%
  filter(coef == "scaled_age" & time == "PreExc") %>%
  group_by(effect) %>%
  mutate(num_targets = n_distinct(target)) %>%
  ungroup() %>%
  mutate(effect_label = paste(effect, "(No of introns:", num_targets, ")"))


# Visualise the plot
aging_plot <- ggplot(Aging_effect, aes(x = scaled_age, y = fit, color = effect_label)) +
  geom_smooth(method = "lm", se = FALSE) + # se = FALSE to remove confidence intervals
  theme_minimal() +
  labs(title = "Relationship between aging and Splicing Efficiency at baseline", 
       x = "Scaled Age", 
       y = "Splicing Efficiency") +
  theme(plot.title = element_text(hjust = 0.5),
        legend.text = element_text(size = 10, face = "italic"),
        legend.title = element_text(size = 10, face = "bold", hjust = 0.5))
print(aging_plot)
```

## Visualize length of introns

This would produce a histogram of the distribution of intron length in the dataset with an inset of the distribution of the length of those affected by aging. This is hoped to gain an insight into how many nucleotide base pairs are likely involved when we refer to ratios or percentage introns retained

```{r}


# Prepare data
length_introns <- Aging_effect %>%
 dplyr::select(target, intron_length, effect) %>%
  distinct()

# Create main plot
main_plot <- length_introns %>%
  ggplot(aes(x = intron_length)) +
  geom_histogram(fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of Intron Lengths",
    x = "Intron Length",
    y = "Number of Introns"
  ) +
  scale_x_continuous(limits = c(70, 30000)) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

# Create inset plot for Reduced SE
Red_plot <- length_introns %>%
  filter(effect == "Reduced SE") %>%
  ggplot(aes(x = intron_length)) +
  geom_histogram(fill = "tomato", color = "white") +
  labs(title = "Declining SE with Aging", x = "Intron Length", y = "Number of Introns") +
  scale_x_continuous(limits = c(70, NA)) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

# Create inset plot for Improved SE
Imp_plot <- length_introns %>%
  filter(effect == "Improved SE") %>%
  ggplot(aes(x = intron_length)) +
  geom_histogram(fill = "seagreen", color = "white") +
  labs(title = "Improving SE with Aging", x = "Intron Length", y = "Number of Introns") +
  scale_x_continuous(limits = c(70, NA)) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

# Combine plots
length_plot <- ggdraw() +
  draw_plot(main_plot) +
  draw_plot(Red_plot, x = 0.65, y = 0.65, width = 0.3, height = 0.3) +
  draw_plot(Imp_plot, x = 0.65, y = 0.25, width = 0.3, height = 0.3)

# Display
print(length_plot)

```

Explore the most affected introns based on their fcthreshold and fit values. It would produce a figure of the actual aggregated SE values per scaled age and the predicted trend line

```{r}
# 1. Extract significantly age-affected introns
doubled_aging <- Aging_effect %>%
  filter(fcthreshold == "s") %>%
  arrange(fit) %>%
  dplyr::select(target, external_gene_name, time, fit, Estimate, transcript_biotype, scaled_age, external_transcript_name)


# 2. Reshape splicing data and join with metadata
long_df <- all_splicing %>%
  pivot_longer(cols = -transcript_ID, names_to = "seq_sample_id", values_to = "SE") %>%
  inner_join(all_metadata, by = "seq_sample_id")

# 3. Select top introns for plotting
top_introns <- c(
  "ENST00000526182.1_1_1", "ENST00000343257.7_20_7", "ENST00000689936.2_70_19",
  "ENST00000258888.6_9_15", "ENST00000504055.1_2_6", "ENST00000487126.5_6_10",
  "ENST00000290219.11_5_21", "ENST00000272167.10_7_1", "ENST00000372642.5_2_9",
  "ENST00000649427.1_5_2", "ENST00000228641.4_2_12", "ENST00000306336.6_3_2"
)

# Predicted fit values
ranked <- Aging_effect %>%
  filter(target %in% top_introns) %>%
  dplyr::select(target, scaled_age, fit) %>%
  mutate(type = "prediction")

# Observed SE values
ranked_original <- long_df %>%
  filter(transcript_ID %in% top_introns, time == "PreExc") %>%
  group_by(scaled_age, transcript_ID) %>%
  summarize(SE = mean(SE), .groups = "drop") %>%
  mutate(target = transcript_ID, fit = SE, type = "SE") %>%
  dplyr::select(target, scaled_age, fit, type)

# 4. Combine and annotate
ranked_df <- bind_rows(ranked, ranked_original) %>%
  mutate(transcript_ID = str_split(target, "_", simplify = TRUE)[, 1]) %>%
  inner_join(gene_annotation, by = c("transcript_ID" = "ensembl_transcript_id_version"))

# 5. Plot
top_decline <- ggplot(ranked_df, aes(x = scaled_age, y = fit, shape = type, colour = type)) +
  geom_point(aes(alpha = 1)) +
  facet_wrap(~target, scales = "free") +
  theme_minimal() +
  scale_alpha_identity() +
  labs(
    title = "Introns with most age-associated decline in SE",
    x = "Scaled Age",
    y = "Splicing Efficiency"
  ) +
  geom_text(
    aes(x = Inf, y = Inf, label = paste("Transcript name:", external_transcript_name)),
    vjust = 1.1, hjust = 1.1, size = 2.5, colour = "darkgrey"
  ) +
  theme(plot.title = element_text(hjust = 0.5)) +
  guides( shape = guide_legend(title = "type"))

print(top_decline)

```

The topmost affected introns appear to be those with declining SE with aging. The code block below will extract and visualize introns with SE that tend to improve with aging

```{r}
# 1. Identify introns with improved SE due to aging
improved_with_aging <- Aging_effect %>%
  filter(effect == "Improved SE", fcthreshold == "s") %>%
  arrange(fit) %>%
  dplyr::select(target, external_gene_name, time, fit, Estimate, transcript_biotype, scaled_age, external_transcript_name)

# 2. Define top improved introns
top_improved_ids <- c(
  "ENST00000369541.4_1_1", "ENST00000304992.11_38_17", "ENST00000640292.2_3_1",
  "ENST00000242728.5_4_12", "ENST00000640575.2_1_2", "ENST00000615631.5_6_8",
  "ENST00000471642.6_8_1", "ENST00000544216.8_4_19", "ENST00000609742.3_4_10",
  "ENST00000558134.5_3_15", "ENST00000456057.5_9_6", "ENST00000650546.1_2_1"
)

# 3. Extract predicted fit values
ranked_improved <- Aging_effect %>%
  filter(target %in% top_improved_ids) %>%
  dplyr::select(target, scaled_age, fit) %>%
  mutate(type = "prediction")

# 4. Extract observed SE values
improved_original <- long_df %>%
  filter(transcript_ID %in% top_improved_ids) %>%
  group_by(scaled_age, transcript_ID) %>%
  summarize(SE = mean(SE), .groups = "drop") %>%
  mutate(target = transcript_ID, fit = SE, type = "SE") %>%
  dplyr::select(target, scaled_age, fit, type)

# 5. Combine and annotate
improved_df <- bind_rows(ranked_improved, improved_original) %>%
  mutate(transcript_ID = str_split(target, "_", simplify = TRUE)[, 1]) %>%
  inner_join(gene_annotation, by = c("transcript_ID" = "ensembl_transcript_id_version"))

# 6. Plot SE improvement with aging
top_improved <- ggplot(improved_df, aes(x = scaled_age, y = fit, shape = type, color = type)) +
  geom_point(alpha = 1) +
  # geom_smooth(method = "lm", se = FALSE, aes(group = type), linetype = "dashed", size = 0.6) +
  facet_wrap(~target, scales = "free") +
  theme_minimal() +
  scale_color_manual(values = c("prediction" = "blue", "SE" = "green")) +
  labs(
    title = "Introns with nost age-Associated Improvement in Splicing Efficiency",
    x = "Scaled Age",
    y = "Splicing Efficiency"
  ) +
  geom_text(
    aes(x = Inf, y = Inf, label = paste("Transcript name:", external_transcript_name)),
    vjust = 1.1, hjust = 1.1, size = 3, colour = "darkgrey"
  ) +
  theme(plot.title = element_text(hjust = 0.5)) +
  guides(shape = guide_legend(title = "Type"), color = guide_legend(title = "Type"))

print(top_improved)
```

Merge the top introns in both categories into one dataframe

```{r}
ggarrange(top_decline,NULL,  top_improved,
          nrow = 3,
          heights = c(1,0.1,1),
          labels = c("A", "", "B"),
          align = "hv")


ggsave("plots/FigureEV2.png", bg = "white",scale = 2.5, dpi = 400)
```

## Gene ontology analysis for genes containing introns affected by aging

```{r}
# select only genes present in the splicing data
gene_exp <- gene_exp_df[gene_exp_df$gene_name %in% RT_merged$external_gene_name, ]

Aging_df <- Aging_effect  %>%
  filter(effect != "No effect")


# Functional annotation of the genes affected
ego_aging <- enrichGO(gene = unique(Aging_df$external_gene_name) ,
                      keyType = "SYMBOL",
                      universe = gene_exp$gene_name,
                      OrgDb = org.Hs.eg.db, 
                      ont = "BP", 
                      pAdjustMethod = "BH", 
                      qvalueCutoff = 0.05, 
                      readable = T)


go_aging <- dotplot(ego_aging,
                    
                    font.size = 8, title = "Enriched biological processes in genes containing introns with aging-associated SE") +
  theme(axis.text = element_text(size = 10), axis.text.y = element_text(size = 8), axis.title.x = element_text(size = 10),
        plot.title = element_text(hjust = 0) )

print(go_aging)
```

Plotting those with improving SE alone didnt produce any results

```{r}
imp_df <- Aging_effect  %>%
  filter(effect == "Improved SE")


# Functional annotation of the genes affected
ego_imp<- enrichGO(gene = unique(imp_df$external_gene_name) ,
                      keyType = "SYMBOL",
                      universe = gene_exp$gene_name,
                      OrgDb = org.Hs.eg.db, 
                      ont = "BP", 
                      pAdjustMethod = "BH", 
                      qvalueCutoff = 0.05, 
                      readable = T)


 dotplot(ego_imp,
                    
                    font.size = 8, title = "Enriched biological processes in genes containing introns with aging-associated improved SE") +
  theme(axis.text = element_text(size = 10), axis.text.y = element_text(size = 8), axis.title.x = element_text(size = 10),
        plot.title = element_text(hjust = 0) )
```

Exploring only those that decline with aging

```{r}
dec_df <- Aging_effect  %>%
  filter(effect == "Reduced SE")


# Functional annotation of the genes affected
ego_dec<- enrichGO(gene = unique(dec_df$external_gene_name) ,
                      keyType = "SYMBOL",
                      universe = gene_exp$gene_name,
                      OrgDb = org.Hs.eg.db, 
                      ont = "BP", 
                      pAdjustMethod = "BH", 
                      qvalueCutoff = 0.05, 
                      readable = T)


 dotplot(ego_dec,
                    
                    font.size = 8, title = "Enriched biological processes in genes containing introns with aging-associated declines in SE") +
  theme(axis.text = element_text(size = 10), axis.text.y = element_text(size = 8), axis.title.x = element_text(size = 10),
        plot.title = element_text(hjust = 0) )
```

### cellular compartments of genes affected by aging

```{r}
# Functional annotation of the genes affected
ego_aging_CC <- enrichGO(gene = unique(Aging_df$external_gene_name) ,
                         keyType = "SYMBOL",
                         universe = gene_exp$gene_name,
                         OrgDb = org.Hs.eg.db, 
                         ont = "CC", 
                         pAdjustMethod = "BH", 
                         qvalueCutoff = 0.05, 
                         readable = T)


go_aging_CC <- dotplot(ego_aging_CC,
                       
                       font.size = 8, title = "Cellular compartment of genes containing introns with aging-associated SE") +
  theme(axis.text = element_text(size = 10), axis.text.y = element_text(size = 10), axis.title.x = element_text(size = 10),
        plot.title = element_text(hjust = 0))

print(go_aging_CC)
```

```{r}

ggarrange(aging_plot, length_plot, go_aging, go_aging_CC,
          ncol = 2, nrow = 2, labels = c("A", "B", "C", "D"),
         #align = "v",
          #          axis = "tblr",
          label.x = 0.05,
          #         label.y = 0.05,
          font.label = list(size = 11),
          heights = c(1.3, 1),
          widths = c(1, 1.2))


ggsave("plots/Figure1.png", bg = "white",scale = 2.5, dpi = 400)
# there was no obvious molecular function
```

The effect of aging on SE was intron-specific. The following script will be visualising that using an upset plot

### Visualise the distribution of the introns in gene biotypes based on the effect of aging on IR

```{r}


# subset these effects into different dataframes
no_effect <- RT_merged %>% 
  filter(effect == "No effect")
Improved_Se <- RT_merged %>%
  filter(effect == "Improved SE")
Reduced_se <- RT_merged %>%
  filter(effect == "Reduced SE")

# Get the intersect based on transcript IDs. This is to determine if there are transcripts
# across the three groups

intersect_df <- Reduce(intersect, list(no_effect$external_gene_name, 
                                       Improved_Se$external_gene_name, Reduced_se$external_gene_name))

intersect_efect <- intersect(no_effect$transcript_ID, Reduced_se$transcript_ID)


# create a Venn list
int_data <- list("No_effect" = no_effect$external_gene_name,
                 "Improved SE" = Improved_Se$external_gene_name,
                 "Reduced SE" = Reduced_se$external_gene_name)

# Plot the UpSet plot
 upset(fromList(int_data), order.by = "freq",
      text.scale = 1.5, 
      mainbar.y.label = "Number of intersecting genes", 
      sets.x.label = "Number of genes", 
      sets.bar.color = c("grey30", "grey50", "grey60"),
      number.angles = 45, 
      point.size = 4.5, 
      matrix.color = "black",
      line.size = 2)


```

## Impact of RT on IR

The impact of RT has those dependent on age, and those independent of age

```{r}
# Explore the effect of aging alone and extract the number of introns per category
Age_dependent_RT <- RT_merged %>%
  filter(coef == "scaled_age:timePostExc") %>%
  group_by(effect) %>%
  mutate(num_targets = n_distinct(target)) %>%
  ungroup() %>%
  mutate(effect_label = paste(effect, "(No of introns:", num_targets, ")"))


# Visualise the plot
age_rt_plot <- ggplot(Age_dependent_RT, aes(x = scaled_age, y = fit, color = effect_label, linetype = time)) +
  geom_smooth(method = "lm", se = FALSE) + # se = FALSE to remove confidence intervals
  theme_minimal() +
  labs(title = "Age-dependent relationship between RT and Splicing Efficiency", 
       x = "Scaled Age", 
       y = "Splicing Efficiency") +
  theme(plot.title = element_text(hjust = 0.5))+
  theme(plot.title = element_text(hjust = 0.5),
        legend.text = element_text(size = 10, face = "italic"),
        legend.title = element_text(size = 12, face = "bold", hjust = 0.5))

age_rt_plot
```

The effects of RT alone, ie the non-age dependent effects

```{r}

# extract only the age-dpendent effects. IE exclude those without effect
Age_dep_RT <- Age_dependent_RT %>%
  filter(effect != "No effect")
# Explore the effect of aging alone and extract the number of introns per category
RT_effect <- RT_merged %>%
  filter(coef == "timePostExc" & effect != "No effect") %>%
  group_by(effect) %>%
  mutate(num_targets = n_distinct(target)) %>%
  ungroup() %>%
  mutate(effect_label = paste(effect, "(No of introns:", num_targets, ")")) %>%
  # To make it exclusive to those not affected by an interaction with age,
  # We exclude those in the age-dependent RT effect group
  filter(!target %in% Age_dep_RT$target)


RT_plot  <- ggplot(RT_effect, aes(x = scaled_age, y = fit, color = effect_label, linetype = time)) +
  geom_smooth(method = "lm", se = FALSE) + # se = FALSE to remove confidence intervals
  theme_minimal() +
  labs(title = "Relationship between  RT and Splicing Efficiency", 
       x = "Scaled Age", 
       y = "Splicing Efficiency") +
  theme(plot.title = element_text(hjust = 0.5))+
  theme(plot.title = element_text(hjust = 0.5),
        legend.text = element_text(size = 10, face = "italic"),
        legend.title = element_text(size = 12, face = "bold", hjust = 0.5))

RT_plot
```

Explore specific introns that are affected by RT

```{r}
RT_ranked <- RT_effect %>%
  filter(time == "PostExc") %>%
  arrange(fit) 

RT_ranked_list <- RT_effect %>%
  filter(target %in% c("ENST00000369541.4_1_1", "ENST00000504055.1_2_6",
                       "ENST00000487126.5_6_10","ENST00000258888.6_9_15",
                       "ENST00000301364.10_13_17","ENST00000714472.1_10_16",
                       "ENST00000553489.1_1_12", "ENST00000228641.4_2_12", 
                       "ENST00000490003.5_1_3", "ENST00000395905.8_6_3",
                       "ENST00000675164.1_16_10", "ENST00000168977.7_6_19"))


Top_affected_introns <- ggplot(RT_ranked_list, aes(x = scaled_age, y = fit, color = time, linetype = time)) +
  geom_smooth(method = "lm", se = FALSE) + # se = FALSE to remove confidence intervals
  theme_minimal() +
  labs(title = "Relationship between RT and Splicing Efficiency in  12  RT_associated introns", 
       x = "Scaled Age", 
       y = "Splicing Efficiency") +
  theme(plot.title = element_text(hjust = 0.5),
        legend.text = element_text(size = 10, face = "italic"),
        legend.title = element_text(size = 12, face = "bold", hjust = 0.5))+
  facet_wrap(~target,scale = "free",  ncol = 3) +
  geom_text(data = RT_ranked_list, aes( x = Inf, y = Inf, label =paste("Transcript name:", external_transcript_name)),
            vjust = 1.1, hjust = 1.1, size = 3, colour = "darkgrey")

Top_affected_introns
```

## GO analysis of the genes containing the RT affected introns

```{r}
# Functional annotation of the genes affected
ego_RT <- enrichGO(gene = unique(RT_effect$external_gene_name),
                   keyType = "SYMBOL",
                   universe = gene_exp_df$gene_name,
                   OrgDb = org.Hs.eg.db, 
                   ont = "BP", 
                   pAdjustMethod = "BH", 
                   qvalueCutoff = 0.05, 
                   readable = T)



go_RT <- dotplot(ego_RT,
                 
                 font.size = 8, title = "Enriched biological processes in genes containing introns with RT-associated SE") +
  theme(axis.text = element_text(size = 10), axis.text.y = element_text(size = 10), axis.title.x = element_text(size = 10))
go_RT




```

\# explore that of those involving RT's interaction with aging

```{r}
# Functional annotation of the genes affected
ego_age_RT <- enrichGO(gene = unique(Age_dep_RT$external_gene_name),
                   keyType = "SYMBOL",
                   universe = gene_exp_df$gene_name,
                   OrgDb = org.Hs.eg.db, 
                   ont = "BP", 
                   pAdjustMethod = "BH", 
                   qvalueCutoff = 0.05, 
                   readable = T)



go_age_RT <- dotplot(ego_age_RT,
                 
                 font.size = 8, title = "Enriched biological processes in genes containing introns with age-dependent RT associations on SE") +
  theme(axis.text = element_text(size = 10), axis.text.y = element_text(size = 10), axis.title.x = element_text(size = 10))
go_RT
go_age_RT
```

```{r}

# Functional annotation of the genes affected
ego_RT_cc <- enrichGO(gene = unique(RT_effect$ensembl_gene_id),
                      keyType = "ENSEMBL",
                      universe = gene_exp_df$gene_id,
                      OrgDb = org.Hs.eg.db, 
                      ont = "cc", 
                      pAdjustMethod = "BH", 
                      qvalueCutoff = 0.05, 
                      readable = T)



go_RT_cc <- dotplot(ego_RT_cc,
                    
                    font.size = 8, title = "Enriched cellular compartments in genes containing introns with RT-associated SE") +
  theme(axis.text = element_text(size = 10), axis.text.y = element_text(size = 10), axis.title.x = element_text(size = 10))

go_RT_cc
```

```{r}
ggarrange(RT_plot, age_rt_plot, go_RT, go_RT_cc,
          labels = c("A", "B", "C", "D"),
          font.label = list(size = 12),
          align = "v")

```
